---
title: "Example Report - State Water Resources Control Board Employment Data"
date: "Updated: `r Sys.Date()`"
output: 
  github_document:
    toc: TRUE
    toc_depth: 5
    # html_preview: TRUE
    # keep_html: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# load packages
library(here)
library(janitor)
library(dplyr)
library(readr)
library(lubridate)
library(ggplot2)
library(forcats)
library(scales)
library(glue)

# enter variables
report_department <- 'Water Resources Control Board'
report_year <- 2020
```

## Introduction

This is an example R Markdown document (in a format used for publishing markdown documents to GitHub - when you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated). 

This example analyzes the demographics of the California State Water Resources Control Board's workforce, using data from [CalHR's Statewide 5102 report](https://www.calhr.ca.gov/pages/statewide-reports.aspx). A cleaned and compiled version of that data is available on the California Open Data Portal, [here](https://data.ca.gov/dataset/calhr-civil-rights-data-for-gare-capital-cohort-2019).

## Including Code

You can include R code in the document as follows:

```{r summary_data}
# import data ---------------
df_5102_report <- read_csv(file = here('03_data_processed',
                                       'calhr_5102_statewide_2011-2020.csv.gz'),
                           col_types = cols(.default = col_character())) %>% 
    type_convert() %>% 
    clean_names()

# add a column for the year
df_5102_report <- df_5102_report %>% 
    mutate(report_year = year(as_of_date))

# view a summary of the dataset
glimpse(df_5102_report)
```

## Including Plots

You can also embed plots, for example (note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot):

```{r plot_department_summary, echo=FALSE}
pl_dept_summary <- df_5102_report %>% 
            filter(dept == report_department,
                   report_year == report_year) %>% 
            group_by(identity_variable, gender) %>% 
            summarize(total_n = sum(record_count)) %>% 
            ungroup() %>% 
            mutate(identity_variable = fct_reorder(identity_variable, total_n)) %>% 
            ggplot() + # code below this line actually creates the plot
            aes(x = identity_variable, y = total_n, fill = gender) +
            geom_bar(stat = 'identity', position = 'dodge') + # position = 'stack'
            scale_y_continuous(labels = comma) +
            coord_flip() +
            labs(x = 'Race / Ethnicity', 
                 y = 'Number of Employees', 
                 title = glue('Employees In California State Government'),
                 subtitle = glue('Year: ', report_year, 
                                 ' | Department: ', report_department), 
                 caption = "Note: data from 5102 Inline Report", 
                 fill = 'Gender')
        
        # output the plot
        pl_dept_summary
```

