---
title: "5112 Data Analysis by Agency / Department"
author: "CalEPA Racial Equity Team"
date: "*Updated: `r format(Sys.Date())`*"
output:
  html_document:
    code_folding: hide
    toc: TRUE
    toc_float: TRUE
    toc_depth: 5
---

# 5112 Data Analysis by Agency / Department

## Introduction
This code can be modified and applied to any department(s) in the California
state system listed in California Department of Human Resource's (CalHR) 5112 data 
set. This code was written by California Environmental Protection Agency's (CalEPA) 
Racial Equity Team (CRET) Data Subteam in Summer 2021 to provide helpful 
racial demographics visualizations for the workforce of the Agency and its Boards, 
Departments, and Office (BDOs). The effort came out of a desire to supplement the
Agency's Race Forward survey results with workforce demographic data. The Data 
Subteam intends on maintaining and building off this and other resources and datasets.
The vision of the Workforce Equity Subteam's work is to create a workforce in which 
one's work and/or career outcomes are not predicted by one's race. The Data Subteam's 
efforts support this vision. This vision is informed by more detailed goals and objectives 
related to work/career satisfaction, promotion and retention, compensation, 
inclusion, mental and other health safety, and more. 
CRET's model of change - normalize, organize and operationalize – is 
aligned with over 200 government organizations part of a nationwide network to 
transform governments to deliver more racial equity. We anticipate that these
demographic metrics and resulting performance measures will evolve over time and 
will be informed by future data and analyses of surveys, workforce demographics, 
and research.
Finally, the Data Subteam acknowledges that racial equity work is complex and 
cannot be pared down to simple numbers in bins. This code and resulting 
visualizations seeks to be one of the first steps in critically examining an agency's 
racial demographics, and may be more helpful in developing the right questions 
that need to be asked, rather than immediate answers and/or numeric targets.
If you have any questions about this code, please email Leah Jones at <leah.jones@waterboards.ca.gov>.

## CA Workforce Data Resources

-   CalHR data page:
    <https://www.calhr.ca.gov/pages/statewide-reports.aspx>

-   Workforce data on CA open data portal:
    <https://data.ca.gov/dataset/calhr-civil-rights-data-for-gare-capital-cohort-2019>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

# load packages ----
library(readxl)
library(readr)
library(dplyr)
library(janitor)
library(here)
library(purrr)
library(glue)
library(lubridate)
library(ggplot2)
library(forcats)
library(scales)
library(tidycensus)
library(sf)
library(stringr)
library(glue)
library(tidyr)
library(patchwork)
library(kableExtra)
```

## Enter Parameters

Enter the year of the most recently published 5102 report and the most
recently published ACS data, and if needed edit the URL for the 5102
data set on the California Open Data Portal. Also, edit default plot
titles and captions as needed (CalEPA and its 6 Boards, Departments, 
and Offices [BDOs] are the examples used here).

```{r}
# enter parameters (for current reporting year) -----------------------------
report_year <- 2020

agency_name <- 'CalEPA'

# pull data from data.ca.gov, this may take a few minutes
# NOTE: change the URL to get data for a different year
df_5112 <- read_csv('https://data.ca.gov/dataset/e620a64f-6b86-4ce0-ab4b-03d06674287b/resource/d308f328-6b5b-41c1-8bc2-a4afcfcee3d1/download/5112-inline-report_12.31.20.csv') %>% 
    clean_names()

url_data_all_yrs <- 'https://data.ca.gov/dataset/e620a64f-6b86-4ce0-ab4b-03d06674287b/resource/aba87ad9-f6b0-4a7e-a45e-d1452417eb7f/download/calhr_5102_statewide_2011-2020.csv'

default_pl_title <- glue('{agency_name} Workforce Compared to State Population (Year {report_year})')
default_pl_title_groups <- glue('{agency_name} Workforce by Occupation and Ethnicity Group Compared to State Population (Year {report_year})')
default_pl_title_bdo <- glue('{agency_name} Workforce by BDO Compared to State Population (Year {report_year})')

default_pl_subtitle <- element_blank()
default_pl_caption <- glue('Data sources: state population data from {acs_data_year} 1 yr American Community Survey  |  workforce data from {report_year} CalHR 5102 Report')
```

## Load & Transform Data

### 5102 Data

CalHR's biannual 5102 report "captures the race/ethnicity, gender, disability, 
and veteran status by occupational group and classification for all state employees."
Data source: [CalHR's Statewide 5102 report](https://www.calhr.ca.gov/pages/statewide-reports.aspx). 
A cleaned and compiled version of that data is available on the California Open Data Portal at: <https://data.ca.gov/dataset/calhr-civil-rights-data-for-gare-capital-cohort-2019>

```{r}
# add additional levels for ethnicity grouping
#   ethnicity (in original dataset) is most detailed level
#   level 1 groups all BIPOC together (resulting groups are BIPOC and white)
#   level 2 groups some ethnicity sub-groupings together (Asian group and Pacific Islander group)
df_5112 <- df_5112 %>% 
    mutate(
        ethnicity_level1 = case_when(
            ethnicity == 'White' ~ 'White', 
            TRUE ~ 'BIPOC'
        ),
        ethnicity_level2 = case_when(
            str_detect(ethnicity, 'Pacific Islander') ~ 'Pacific Islander', # groups (all start with "Pacific Islander - "): Guamanian, Hawaiian, Other or Multiple, and Samoan
            str_detect(ethnicity, 'Asian') ~ 'Asian', # groups (all start with "Asian - "): Cambodian, Chinese, Filipino, Indian, Japanese, Korean, Laotian, Other or Multiple, Vietnamese
            TRUE ~ ethnicity)
    )

# Filtering for EPA and related BDOs 
df_5112_epa <- df_5112 %>% 
    filter(department == "Air Resources Board"|
               department == "Environmental Health Hazard Assessment, Office of"|
               department == "Environmental Protection Agency"|
               department == "Pesticide Regulation, Department of"|
               department == "Resources Recycling and Recovery, Department of"|
               department == "Toxic Substances Control, Department of"|
               department == "Water Resources Control Board")
```



    

```{r}
# 2 - 5112 exploration ----------------------------------------------------------------
## count of entries by race ----
ggplot() +
    geom_bar(data = df_5112_epa, 
             aes(ethnicity)) +
    coord_flip()



# 3 - 5112 All-CalEPA rates ----------------------------------------------------------

## define color palette for rate plots
# RColorBrewer::brewer.pal(n = 3, name = 'Paired')
colors_rate_plots <- c('darkblue', 'lightblue', 'gold') # intake outside, intake other agencies, advancements

### detailed ethnic groups ----
# original ethnicity 
sum_5112_detail <- df_5112_epa %>% 
    select(ethnicity, hire_type) %>% 
    add_count(ethnicity, name = 'ethnicity_total') %>% 
    add_count(ethnicity, hire_type, name = 'ethnicity_type_total') %>% 
    mutate(rate = ethnicity_type_total / ethnicity_total) %>% 
    distinct() %>%
    arrange(ethnicity) %>% 
    mutate(data_label = glue('{ethnicity} \n(n = {ethnicity_total})')) %>% 
    mutate(hire_type = factor(hire_type)) %>% 
    # stuff below here is for adding labels to each segment of the bars in the bar charts
    mutate(hire_type = as.factor(hire_type)) %>%  
    mutate(hire_type = fct_relevel(hire_type, 
                                   c('Advancements', 
                                     'Intake: Hires from other State Agencies', 
                                     'Intake: Outside Hires'))) %>% 
    arrange(ethnicity, hire_type) %>% 
    group_by(ethnicity) %>% 
    mutate(label_y = cumsum(rate) - 0.5 * rate) %>% # puts the labels in the middle of each bar
    ungroup() %>%
    {.}

# create ordering for bars (by number of advancements)
ordering_detail <- sum_5112_detail %>%
    filter(hire_type == 'Advancements') %>%
    arrange(desc(rate)) %>%
    pull(ethnicity) %>%
    as.character()
missing_detail <- unique(sum_5112_detail$ethnicity)[!unique(sum_5112_detail$ethnicity) %in% ordering_detail]
ordering_detail <- c(ordering_detail, missing_detail)
# ordering <- sum_5112_detail %>% 
#     filter(hire_type == 'Intake: Outside Hires') %>% 
#     arrange(desc(rate)) %>% 
#     pull(ethnicity) %>% 
#     as.character()
# ordering_2 <- sum_5112_detail %>% 
#     filter(hire_type == 'Intake: Hires from other State Agencies') %>% 
#     arrange(desc(rate)) %>% 
#     pull(ethnicity) %>% 
#     as.character()
# missing_1 <- ordering_2[!ordering_2 %in% ordering]
# ordering <- c(ordering, missing_1)
# missing_2 <- unique(sum_5112_detail$ethnicity)[!unique(sum_5112_detail$ethnicity) %in% ordering]
# ordering <- c(ordering, missing_2)
data_labels_detail <- tibble(ethnicity = ordering_detail) %>%
    left_join(sum_5112_detail %>% distinct(ethnicity, data_label)) %>%
    pull(data_label)

(pl_5112_detail <- sum_5112_detail %>% 
        mutate(ethnicity = as.factor(ethnicity)) %>% 
        mutate(ethnicity = fct_relevel(ethnicity, rev(ordering_detail))) %>%
        ggplot() +
        geom_bar(mapping = aes(x = ethnicity, # data_label, 
                               y = rate, 
                               fill = fct_rev(hire_type)), 
                 stat = 'identity') +
        scale_fill_manual(values = colors_rate_plots) +
        scale_y_continuous(labels = percent) +
        scale_x_discrete(breaks = ordering_detail,
                         labels = data_labels_detail) +
        labs(title = glue('Intake Versus Advancement by Ethnicity for all CalEPA BDOs (Year {report_year})'),
             subtitle = 'Labels represent total number in each grouping',
             x = 'Ethnicity Group',
             y = 'Percent of Total Intakes and Advancements (by Ethnicity Group)', 
             caption = glue('Data Source: {report_year} CalHR 5112 Report')) +
        geom_text(aes(x = ethnicity, # data_label, # 
                      y = label_y,
                      # label = ethnicity_type_total
                      label = glue('{ethnicity_type_total}')
                      ),
                  color = 'grey50') +
        coord_flip() + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        guides(fill = guide_legend(reverse = TRUE)))

ggsave(filename = here('07_slides', '2021-05-14', 'images', '5112_calepa_detail.png'), 
       plot = pl_5112_detail, 
       width = 10, 
       height = 6, 
       dpi = 125
       )



### L1 ethnicity ----
# level 1 ethnicity (white vs BIPOC)
sum_5112_l1 <- df_5112_epa %>% 
    select(ethnicity_level1, hire_type) %>% 
    add_count(ethnicity_level1, name = 'ethnicity_total') %>% 
    add_count(ethnicity_level1, hire_type, name = 'ethnicity_type_total') %>% 
    distinct() %>% 
    mutate(rate = ethnicity_type_total / ethnicity_total) %>% 
    arrange(ethnicity_level1) %>%
    mutate(data_label = glue('{ethnicity_level1} \n(n = {ethnicity_total})')) %>% 
    # stuff below here is for adding labels to each segment of the bars in the bar charts
    mutate(hire_type = as.factor(hire_type)) %>%  
    mutate(hire_type = fct_relevel(hire_type, 
                                   c('Advancements',
                                     'Intake: Hires from other State Agencies', 
                                     'Intake: Outside Hires'))) %>% 
    arrange(ethnicity_level1, hire_type) %>% 
    group_by(ethnicity_level1) %>% 
    mutate(label_y = cumsum(rate) - 0.5 * rate) %>% # puts the labels in the middle of each bar
    ungroup() %>%
    {.}

(pl_5112_l1 <- sum_5112_l1 %>%
        # mutate(data_label = glue('{ethnicity_level1} \n(n = {ethnicity_total})')) %>% 
        mutate(hire_type = fct_rev(hire_type)) %>% 
        ggplot() +
        geom_bar(mapping = aes(x = data_label, 
                               #x = ethnicity_level1,
                               y = rate, 
                               fill = hire_type), 
                 stat = 'identity') +
        scale_y_continuous(labels = percent) +
        labs(title = glue('Intake Versus Advancement by Ethnicity for all CalEPA BDOs (Year {report_year})'),
             subtitle = 'Labels represent total number in each grouping',
             x = 'Ethnicity Group',
             y = 'Percent of Total Intakes and Advancements (by Ethnicity Group)', 
             caption = glue('Data Source: {report_year} CalHR 5112 Report')) +
        guides(fill = guide_legend(reverse = TRUE)) +
        scale_fill_manual(values = colors_rate_plots) + 
        # geom_text(aes(x = ethnicity_level1,
        #               y = 1.05,
        #               label = glue('n = {ethnicity_total}'),
        #               hjust = 0.5)) +
        geom_text(aes(x = data_label, # ethnicity_level1,
                      y = label_y,
                      # label = ethnicity_type_total,
                      label = glue('{ethnicity_type_total}')),
                  color = 'grey50') +
        coord_flip() + 
        # facet_wrap(~ metrics_group, scales = 'free_y') + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        geom_blank())

ggsave(filename = here('07_slides', '2021-05-14', 'images', '5112_calepa_level1.png'), 
       plot = pl_5112_l1, 
       width = 10, 
       height = 6, 
       dpi = 125
       )

### L2 ethnicity ----
# level 2 ethnicity (white vs BIPOC)
sum_5112_l2 <- df_5112_epa %>% 
    select(ethnicity_level2, hire_type) %>% 
    add_count(ethnicity_level2, name = 'ethnicity_total') %>% 
    add_count(ethnicity_level2, hire_type, name = 'ethnicity_type_total') %>% 
    distinct() %>% 
    mutate(rate = ethnicity_type_total / ethnicity_total) %>% 
    arrange(ethnicity_level2) %>%
    mutate(data_label = glue('{ethnicity_level2} \n(n = {ethnicity_total})')) %>% 
    # stuff below here is for adding labels to each segment of the bars in the bar charts
    mutate(hire_type = as.factor(hire_type)) %>%  
    mutate(hire_type = fct_relevel(hire_type, 
                                   c('Advancements',
                                     'Intake: Hires from other State Agencies', 
                                     'Intake: Outside Hires'))) %>% 
    arrange(ethnicity_level2, hire_type) %>% 
    group_by(ethnicity_level2) %>% 
    mutate(label_y = cumsum(rate) - 0.5 * rate) %>% # puts the labels in the middle of each bar
    ungroup() %>%
    {.}

ordering_l2 <- sum_5112_l2 %>%
    filter(hire_type == 'Advancements') %>%
    arrange(desc(rate)) %>%
    pull(ethnicity_level2) %>%
    as.character()
missing_l2 <- unique(sum_5112_l2$ethnicity_level2)[!unique(sum_5112_l2$ethnicity_level2) %in% ordering_l2]
ordering_l2 <- c(ordering_l2, missing_l2)
data_labels_l2 <- tibble(ethnicity_level2 = ordering_l2) %>%
    left_join(sum_5112_l2 %>% distinct(ethnicity_level2, data_label)) %>%
    pull(data_label)

(pl_5112_l2 <- sum_5112_l2 %>%
        # mutate(data_label = glue('{ethnicity_level1} \n(n = {ethnicity_total})')) %>% 
        mutate(hire_type = fct_rev(hire_type)) %>% 
        mutate(ethnicity_level2 = as.factor(ethnicity_level2)) %>% 
        mutate(ethnicity_level2 = fct_relevel(ethnicity_level2, rev(ordering_l2))) %>%
        ggplot() +
        geom_bar(mapping = aes(# x = data_label, 
                               x = ethnicity_level2,
                               y = rate, 
                               fill = hire_type), 
                 stat = 'identity') +
        scale_y_continuous(labels = percent) +
        labs(title = glue('Intake Versus Advancement by Ethnicity for all CalEPA BDOs (Year {report_year})'),
             subtitle = 'Labels represent total number in each grouping',
             x = 'Ethnicity Group',
             y = 'Percent of Total Intakes and Advancements (by Ethnicity Group)', 
             caption = glue('Data Source: {report_year} CalHR 5112 Report')) +
        guides(fill = guide_legend(reverse = TRUE)) +
        scale_fill_manual(values = colors_rate_plots) + 
        scale_x_discrete(breaks = ordering_l2,
                         labels = data_labels_l2) +
        # geom_text(aes(x = ethnicity_level1,
        #               y = 1.05,
        #               label = glue('n = {ethnicity_total}'),
        #               hjust = 0.5)) +
        geom_text(aes(x = ethnicity_level2,
                      y = label_y,
                      # label = ethnicity_type_total,
                      label = glue('{ethnicity_type_total}')),
                  color = 'grey50') +
        coord_flip() + 
        # facet_wrap(~ metrics_group, scales = 'free_y') + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        geom_blank())

ggsave(filename = here('07_slides', '2021-05-14', 'images', '5112_calepa_level2.png'), 
       plot = pl_5112_l2, 
       width = 10, 
       height = 6, 
       dpi = 125
       )

# patchwork (combine plots)
pl_5512_patch <- pl_5112_l1 / pl_5112_l2 
pl_5512_patch[[1]] <- pl_5512_patch[[1]] + 
    labs(title = element_blank(),
         subtitle = element_blank(),
         caption = element_blank()) + 
    theme(legend.position = 'none')
pl_5512_patch[[2]] <- pl_5512_patch[[2]] + 
    labs(title = element_blank(),
         subtitle = element_blank(),
         caption = element_blank())
pl_5512_patch <- pl_5512_patch + 
    plot_layout(heights = c(1, 3)) +
    plot_annotation(
        title = glue('Intake Versus Advancement by Ethnicity for all CalEPA BDOs (Year {report_year})'),
        subtitle = 'Labels represent total number in each grouping',
        caption = glue('Data Source: {report_year} CalHR 5112 Report'))

ggsave(filename = here('07_slides', '2021-05-14', 'images', '5112_calepa_level1_2_combined.png'), 
       plot = pl_5512_patch, 
       width = 10, 
       height = 6, 
       dpi = 125
       )

# 4 - 5112 group summaries ------------------------------------------------------
## create groupings ----
df_5112_epa <- df_5112_epa %>% 
    mutate(metrics_group = case_when(
        # legal
        soc_major_group_title == 'Legal Occupations' ~ 
            'Legal Occupations', 
        # management
        soc_major_group_title == 'Management Occupations' ~ 
            'Management Occupations', 
        # admin
        soc_major_group_title == 'Office and Administrative Support Occupations' |
            (soc_major_group_title == 'Computer and Mathematical Occupations' & 
                 soc_detailed_group_title == 'Unmapped Classes') ~ 
            'Administrative Occupations',
        # technical
        soc_major_group_title == 'Architecture and Engineering Occupations' |
            soc_major_group_title == 'Life, Physical, and Social Science Occupations' | 
            (soc_major_group_title == 'Computer and Mathematical Occupations' & 
                 soc_detailed_group_title == 'Operations Research Analysts') ~ 
            'Technical Occupations')
    )

## L1 summary by group ----
sum_5112_groups_l1 <- df_5112_epa %>% 
    filter(!is.na(metrics_group)) %>% 
    select(ethnicity_level1, hire_type, metrics_group) %>% 
    add_count(ethnicity_level1, metrics_group, name = 'ethnicity_total') %>% 
    add_count(ethnicity_level1, hire_type, metrics_group, name = 'ethnicity_type_total') %>% 
    add_count(metrics_group, name = 'metrics_group_total') %>% 
    distinct() %>% 
    mutate(rate = ethnicity_type_total / ethnicity_total) %>% 
    # stuff below here is for adding labels to each segment of the bars in the bar charts
    mutate(hire_type = as.factor(hire_type)) %>%  
    mutate(hire_type = fct_relevel(hire_type, 
                                   c('Advancements',
                                     'Intake: Outside Hires', 
                                     'Intake: Hires from other State Agencies'))) %>% 
    arrange(ethnicity_level1, metrics_group, hire_type) %>% 
    group_by(ethnicity_level1, metrics_group) %>% 
    mutate(label_y = cumsum(rate) - 0.5 * rate) %>% # puts the labels in the middle of each bar
    ungroup() %>%
    {.}

(pl_5112_groups_l1 <- sum_5112_groups_l1 %>%
        mutate(data_label = glue('{ethnicity_level1} \n(n = {ethnicity_total})')) %>% 
        mutate(facet_label = glue('{metrics_group} (n = {metrics_group_total})')) %>% 
        mutate(hire_type = fct_rev(hire_type)) %>% 
        # mutate(ethnicity_level1 = factor(ethnicity_level1)) %>% 
        # mutate(ethnicity_level1 = fct_rev(ethnicity_level1)) %>% 
        # mutate(data_label = factor(data_label)) %>% 
        # mutate(data_label = fct_rev(data_label)) %>%
        ggplot() +
        geom_bar(mapping = aes(x = data_label, 
                               # x = ethnicity_level1,
                               y = rate, 
                               fill = hire_type), 
                 stat = 'identity') +
        scale_y_continuous(labels = percent) +
        labs(title = glue('Intake Versus Advancement by Occupation and Ethnicity for all CalEPA BDOs (Year {report_year})'),
             subtitle = 'Labels represent total number in each grouping',
             x = 'Ethnicity Group',
             y = 'Percent of Total Intakes and Advancements (by Occupation and Ethnicity Group)', 
             caption = glue('Data Source: {report_year} CalHR 5112 Report')) +
        guides(fill = guide_legend(reverse = TRUE)) +
        scale_fill_manual(values = colors_rate_plots) + 
        # geom_text(aes(x = ethnicity_level1,
        #               y = 1.05,
        #               label = glue('n = {ethnicity_total}'),
        #               hjust = 0.5)) +
        geom_text(aes(x = data_label, # ethnicity_level1,
                      y = label_y,
                      # label = ethnicity_type_total,
                      label = glue('{ethnicity_type_total}')),
                  color = 'grey50') +
        coord_flip() + 
        facet_wrap(~ facet_label, scales = 'free_y') + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        geom_blank())

ggsave(filename = here('07_slides', '2021-05-14', 'images', 
                       '5112_calepa_groups_level1.png'), 
       plot = pl_5112_groups_l1, 
       width = 10, 
       height = 6, 
       dpi = 125
       )
```       