---
title: "5102 Data Analysis by Agency / Department"
author: "CalEPA Racial Equity Workgroup"
date: "*Updated: `r format(Sys.Date())`*"
output:
  html_document:
    code_folding: hide
    toc: TRUE
    toc_float: TRUE
    toc_depth: 5
---

# 5102 Data Analysis by Agency / Department

## Introduction

intro text...

## Resources

### Census Data:

-   tidycensus R package:
    <https://walker-data.com/tidycensus/articles/basic-usage.html>

-   general census workshop w/ R:
    <https://jennhuck.github.io/workshops/tidycensus.html>

-   ACS Table B02001 data (race w/o hispanic/latino data):
    <https://censusreporter.org/data/table/?table=B02001&geo_ids=04000US06>

-   ACS Table B03002 data (race with hispanic/latino data):
    <https://censusreporter.org/data/table/?table=B03002&geo_ids=04000US06>

-   info about race in census data:
    <https://censusreporter.org/topics/race-hispanic/>

-   other general info about concepts and tables in census data:
    <https://censusreporter.org/topics/>

-   check/validation of the numbers for CA:
    <https://www.ppic.org/publication/californias-population/>

### CA Workforce Data:

-   CalHR data page:
    <https://www.calhr.ca.gov/pages/statewide-reports.aspx>

-   Workforce data on CA open data portal:
    <https://data.ca.gov/dataset/calhr-civil-rights-data-for-gare-capital-cohort-2019>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load packages ----
library(readxl)
library(readr)
library(dplyr)
library(janitor)
library(here)
library(purrr)
library(glue)
library(lubridate)
library(ggplot2)
library(forcats)
library(scales)
library(tidycensus)
library(sf)
library(stringr)
library(glue)
library(tidyr)
library(patchwork)
library(kableExtra)
```

## Enter Parameters

Enter the year of the most recently published 5102 report and the most
recently published ACS data, and if needed edit the URL for the 5102
dataset on the California Open Data Portal. Also, edit default plot
titles and captions as needed.

```{r}
# enter parameters (for current reporting year) -----------------------------
report_year <- 2020
acs_data_year <- 2019

agency_name <- 'CalEPA'

url_data_all_yrs <- 'https://data.ca.gov/dataset/e620a64f-6b86-4ce0-ab4b-03d06674287b/resource/aba87ad9-f6b0-4a7e-a45e-d1452417eb7f/download/calhr_5102_statewide_2011-2020.csv'

default_pl_title <- glue('{agency_name} Workforce Compared to State Population (Year {report_year})')
default_pl_title_groups <- glue('{agency_name} Workforce by Occupation and Ethnicity Group Compared to State Population (Year {report_year})')
default_pl_title_bdo <- glue('{agency_name} Workforce by BDO Compared to State Population (Year {report_year})')

default_pl_subtitle <- element_blank()
default_pl_caption <- glue('Data sources: state population data from {acs_data_year} 1 yr American Community Survey  |  workforce data from {report_year} CalHR 5102 Report')
```

## Load & Transform Data

### 5102 Data

Info about 5102 dataset?...

#### Ethnicity Groupings

describe ethnicity levels added to the 5102 dataset (level 1: BIPOC /
white, level 2: combined Asian and Pacific Islander groupings)...

```{r}
# 1 - load / transform data ---------------------------------------------------------------

## 5102 ----
df_5102 <- read_csv(url_data_all_yrs) %>%
    clean_names() 

# add additional levels for ethnicity grouping
#   identity_variable (in original dataset) is most detailed level
#   level 1 groups all BIPOC together (resulting groups are BIPOC and white)
#   level 2 groups some identity_variable sub-groupings together (Asian group and Pacific Islander group)
df_5102 <- df_5102 %>% 
    mutate(
        ethnicity_level1 = case_when(
            identity_variable == 'White' ~ 'White', 
            TRUE ~ 'BIPOC'),
        ethnicity_level2 = case_when(
            str_detect(identity_variable, 'Pacific Islander') ~ 'Pacific Islander', # groups (all start with "Pacific Islander - "): Guamanian, Hawaiian, Other or Multiple, and Samoan
            str_detect(identity_variable, 'Asian') ~ 'Asian', # groups (all start with "Asian - "): Cambodian, Chinese, Filipino, Indian, Japanese, Korean, Laotian, Other or Multiple, Vietnamese
            TRUE ~ identity_variable)
    )

# add a year column
df_5102 <- df_5102 %>% 
    mutate(reporting_year = year(as_of_date))

```

#### Filter for Selected Agency

List of agency names in the 5102 dataset:
```{r}
# list of all departments
depts_list <- df_5102 %>% 
    distinct(dept) %>% 
    arrange(dept) %>% 
    rename('Department Name' = dept)


depts_list %>% 
    kbl() %>% 
    kable_paper() %>%
    scroll_box(width = "500px", height = "200px") %>%
    {.}
```


Enter departments to be included in the analysis:

```{r}
# Filtering for EPA and related BDOs 
df_5102_epa <- df_5102 %>% 
    filter(dept == "Air Resources Board"|
               dept == "Environmental Health Hazard Assessment, Office of"|
               dept == "Environmental Protection Agency"|
               dept == "Pesticide Regulation, Department of"|
               dept == "Resources Recycling and Recovery, Department of"|
               dept == "Toxic Substances Control, Department of"|
               dept == "Water Resources Control Board")

# filter for data from the report year (selected above)
df_5102_epa_1yr <- df_5102_epa %>% 
    filter(reporting_year == report_year)

```

### Census Data

explain ethnicity levels available in the data...

#### Census Data Resources

-   ACS Table B02001 data (race w/o hispanic/latino data):
    <https://censusreporter.org/data/table/?table=B02001&geo_ids=04000US06>
-   ACS Table B03002 data (race with hispanic/latino data):
    <https://censusreporter.org/data/table/?table=B03002&geo_ids=04000US06>
-   info about race in census data:
    <https://censusreporter.org/topics/race-hispanic/>

#### Download Census Data

explain selected table / variables?...

```{r}
## census data (ACS 1 yr) ----
### view all variables
# acs_variables <- load_variables(acs_data_year, 'acs1', cache = TRUE)
    # to view label for a given code (name)
    # acs_variables %>% filter(name == 'B03002_009') %>% pull(label)

### get ACS data
acs_data_raw <- get_acs(geography = 'state', 
                        variables = c(# total_state_pop = 'B02001_001',
                            'Hispanic or Latino' = 'B03002_012', # Total Hispanic or Latino
                            'White' = 'B03002_003', # White (Not Hispanic or Latino)
                            'Black or African American' = 'B03002_004', # Black or African American (Not Hispanic or Latino)
                            'Native American or Alaska Native' = 'B03002_005', # American Indian and Alaska Native (Not Hispanic or Latino)
                            'Asian' = 'B03002_006', # Asian (Not Hispanic or Latino)
                            'Pacific Islander' = 'B03002_007', # Native Hawaiian and Other Pacific Islander (Not Hispanic or Latino)
                            'Other' = 'B03002_008', # Some other race (Not Hispanic or Latino)
                            'Multiple' = 'B03002_009'# Two or more races (Not Hispanic or Latino)
                        ),
                        summary_var = c(total_state_pop = 'B02001_001'), 
                        survey = 'acs1', # use 'acs1' or 'acs5' to get 1 or 5 year acs
                        state = 'CA', 
                        geometry = FALSE, # set to TRUE to get as geospatial data
                        year = acs_data_year)

# clean / reformat the acs data (the data will be formatted to be 
# consistent with the "level 2" ethnicity groupings in the workforce 
# data that are created above)
acs_data_level2 <- acs_data_raw %>% 
    clean_names() %>% 
    rename(total_state_pop = summary_est, 
           total_state_pop_moe = summary_moe,
           location_name = name,
           ethnicity_level2 = variable)

    ## check (should be TRUE) - make sure sum of populations by group == total state population
    # sum(acs_data_level2$estimate) == acs_data_level2$total_state_pop[1]

### group the 'Other' and 'Multiple' rows into one 'Other or Multiple Race' row
acs_data_level2 <- acs_data_level2 %>% 
    mutate(ethnicity_level2 = 
               case_when(ethnicity_level2 == 'Other' | 
                             ethnicity_level2 == 'Multiple' ~ 
                             'Other or Multiple Race',
                         TRUE ~ ethnicity_level2)) %>% 
    group_by(geoid, location_name, ethnicity_level2, total_state_pop) %>% 
    summarize(estimate = sum(estimate)) %>%
    ungroup() %>% 
    {.}
    
    ## check (should be TRUE) - make sure sum of populations by group == total state population
    # sum(acs_data_level2$estimate) == acs_data_level2$total_state_pop[1]
    
### add a column with each ethnicity's % of total state population
acs_data_level2 <- acs_data_level2 %>% 
    mutate(pop_pct = estimate / total_state_pop) %>% 
    {.}

### create a dataset grouped at level 1 - all BIPOC together (resulting groups are BIPOC and white)
acs_data_level1 <- acs_data_level2 %>%  
    mutate(
        ethnicity_level1 = case_when(
            ethnicity_level2 == 'White' ~ 'White', 
            TRUE ~ 'BIPOC')
    ) %>% 
    group_by(geoid, location_name, ethnicity_level1, total_state_pop) %>% 
    summarize(estimate = sum(estimate)) %>% 
    ungroup() %>% 
    mutate(pop_pct = estimate / total_state_pop) %>% # update the pop_pct to reflect level 1 numbers
    {.}
    
    ## check 
    # sum(acs_data_level1$estimate) == acs_data_level1$total_state_pop[1]
```

## Plots

### Enter Plotting Parameters
```{r}
# !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
# !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
# 2 - 5102 summaries (vs CA pop) -----------------------------------------------
## define color palette for plots
### state pop vs CalEPA workforce plots
colors_5102_state_epa <-c('gold', 'darkgreen') # state pop / CalEPA workforce
### colors for ethnicities (in stacked bar plots) - trying to keep 'white' the same for both levels of plots for consistency/comparison
brewer_colors <- rev(RColorBrewer::brewer.pal(n = 7, name = 'Set2'))
colors_5102_2 <- brewer_colors[6:7] # BIPOC / White
colors_5102_7 <- brewer_colors # detailed ethnic groups: Asian ... White
    

```


### Format Data & Create Plotting Functions

```{r}

## format census data for plot (all of CA) ----
acs_data_level1 <- acs_data_level1 %>%
    mutate(type = factor('State Population')) %>%
    rename(ethnicity_total = estimate,
           rate = pop_pct) %>%
    
    {.}

acs_data_level2 <- acs_data_level2 %>%
    mutate(type = factor('State Population')) %>%
    rename(ethnicity_total = estimate,
           rate = pop_pct) %>%
    {.}

## 5102 summary functions
fun_summary_5102_l1 <- function(dataset) {
    dataset %>%
        add_count(ethnicity_level1,
                  wt = record_count,
                  name = 'ethnicity_total') %>%
        select(ethnicity_level1, ethnicity_total) %>%
        distinct() %>%
        mutate(rate = ethnicity_total / sum(ethnicity_total)) %>%
        mutate(type = 'CalEPA Workforce') %>%
        arrange(ethnicity_level1) %>%
        mutate(type = factor(type)) %>% 
        {.}
}

fun_summary_5102_l2 <- function(dataset) {
    dataset %>%
        add_count(ethnicity_level2,
                  wt = record_count,
                  name = 'ethnicity_total') %>%
        select(ethnicity_level2, ethnicity_total) %>%
        distinct() %>%
        mutate(rate = ethnicity_total / sum(ethnicity_total)) %>%
        mutate(type = 'CalEPA Workforce') %>%
        arrange(ethnicity_level2) %>%
        mutate(type = factor(type)) %>% 
        {.}
}

## plot function
fun_plot_5102_l1 <- function(dataset, plot_title, plot_caption) {
    dataset %>% 
        ggplot() +
        geom_bar(mapping = aes(x = ethnicity_level1, 
                               y = rate, 
                               fill = factor(type, levels = rev(levels(type)))), 
                 stat = 'identity', 
                 position = 'dodge') +
        scale_fill_manual(values = colors_5102_state_epa) +
        scale_y_continuous(labels = label_percent(accuracy = 1L)) +
        labs(title = plot_title,
             x = 'Ethnicity Group',
             y = 'Percent of Total', 
             caption = plot_caption) +
        coord_flip() + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        guides(fill = guide_legend(reverse = TRUE))
}

fun_plot_5102_l1_stack <- function(dataset, plot_title, plot_caption) {
    dataset %>% 
        mutate(type = fct_rev(type)) %>% 
        ggplot() +
        geom_bar(mapping = aes(x = type, # ethnicity_level1, 
                               y = rate, 
                               fill = ethnicity_level1), # factor(type, levels = rev(levels(type)))), 
                 stat = 'identity') +
        scale_fill_manual(values = colors_5102_2) +
        scale_y_continuous(labels = label_percent(accuracy = 1L)) +
        labs(title = plot_title,
             x = element_blank(),
             y = 'Percent of Total', 
             caption = plot_caption) + 
        coord_flip() + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        guides(fill = guide_legend(reverse = TRUE))
}


fun_plot_5102_l2 <- function(dataset, plot_title, plot_caption) {
    dataset %>% 
        ggplot() +
        geom_bar(mapping = aes(x = ethnicity_level2, 
                               y = rate, 
                               fill = factor(type, levels = rev(levels(type)))), 
                 stat = 'identity', 
                 position = 'dodge') +
        scale_fill_manual(values = colors_5102_state_epa) +
        scale_y_continuous(labels = label_percent(accuracy = 1L)) +
        labs(title = plot_title,
             x = 'Ethnicity Group',
             y = 'Percent of Total', 
             caption = plot_caption) +
        coord_flip() + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        guides(fill = guide_legend(reverse = TRUE))
}

fun_plot_5102_l2_stack <- function(dataset, plot_title, plot_caption) {
    dataset %>% 
        mutate(type = fct_rev(type)) %>%
        ggplot() +
        geom_bar(mapping = aes(x = type, 
                               y = rate, 
                               fill = ethnicity_level2), # factor(type, levels = rev(levels(type)))), 
                 stat = 'identity') +
        scale_fill_manual(values = colors_5102_7) +
        scale_y_continuous(labels = label_percent(accuracy = 1L)) +
        labs(title = plot_title,
             x = element_blank(),
             y = 'Percent of Total', 
             caption = plot_caption) +
        coord_flip() + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        guides(fill = guide_legend(reverse = TRUE))
}
```

### Plots...

```{r}
## 2.1 - CalEPA summary ----
### L1 ----
(pl_5102_calepa_l1 <- df_5102_epa_1yr %>% 
     fun_summary_5102_l1() %>% 
     bind_rows(acs_data_level1) %>% 
     fun_plot_5102_l1(plot_title = default_pl_title, 
                      plot_caption = default_pl_caption))

(pl_5102_calepa_l1_stack <- df_5102_epa_1yr %>% 
        fun_summary_5102_l1() %>% 
        bind_rows(acs_data_level1) %>% 
        fun_plot_5102_l1_stack(plot_title = default_pl_title, 
                               plot_caption = default_pl_caption))

ggsave(filename = here('07_slides', '2021-05-14', 'images', 
                       '5102_calepa_level1_stack.png'), 
       plot = pl_5102_calepa_l1_stack, 
       width = 10, 
       height = 6, 
       dpi = 125
       )

# # patchwork (combine plots) - L1
# pl_5512_patch_calepa_l1 <- pl_5102_calepa_l1_stack / pl_5102_calepa_l1

### L2 ----
(pl_5102_calepa_l2 <- df_5102_epa_1yr %>% 
     fun_summary_5102_l2() %>% 
     bind_rows(acs_data_level2) %>% 
     fun_plot_5102_l2(plot_title = default_pl_title, 
                      plot_caption = default_pl_caption))
 
ggsave(filename = here('07_slides', '2021-05-14', 'images', 
                       '5102_calepa_level2.png'), 
       plot = pl_5102_calepa_l2, 
       width = 10, 
       height = 6, 
       dpi = 125
       )

(pl_5102_calepa_l2_stack <- df_5102_epa_1yr %>% 
        fun_summary_5102_l2() %>% 
        bind_rows(acs_data_level2) %>% 
        fun_plot_5102_l2_stack(plot_title = default_pl_title, 
                               plot_caption = default_pl_caption))

ggsave(filename = here('07_slides', '2021-05-14', 'images', 
                       '5102_calepa_level2_stack.png'), 
       plot = pl_5102_calepa_l2_stack, 
       width = 10, 
       height = 6, 
       dpi = 125
       )

# patchwork (combine plots) - L1 and L2 stacked plots
pl_5102_patch_calepa_stack_combined <- pl_5102_calepa_l1_stack / pl_5102_calepa_l2_stack
pl_5102_patch_calepa_stack_combined[[1]] <- pl_5102_patch_calepa_stack_combined[[1]] + 
    labs(title = element_blank(),
         subtitle = element_blank(),
         caption = element_blank())
pl_5102_patch_calepa_stack_combined[[2]] <- pl_5102_patch_calepa_stack_combined[[2]] + 
    labs(title = element_blank(),
         subtitle = element_blank(),
         caption = element_blank())
pl_5102_patch_calepa_stack_combined <- pl_5102_patch_calepa_stack_combined + 
    plot_layout(heights = c(1.7, 2)) +
    plot_annotation(
        title = default_pl_title,
        caption = default_pl_caption)
ggsave(filename = here('07_slides', '2021-05-14', 'images', 
                       '5102_calepa_level1_2_combined_stack.png'), 
       plot = pl_5102_patch_calepa_stack_combined, 
       width = 10, 
       height = 6, 
       dpi = 125
       )

# patchwork (combine plots) - both L2 plots
pl_5102_patch_calepa_l2 <- pl_5102_calepa_l2_stack / pl_5102_calepa_l2
pl_5102_patch_calepa_l2[[1]] <- pl_5102_patch_calepa_l2[[1]] + 
    labs(title = element_blank(),
         subtitle = element_blank(),
         caption = element_blank())
pl_5102_patch_calepa_l2[[2]] <- pl_5102_patch_calepa_l2[[2]] + 
    labs(title = element_blank(),
         subtitle = element_blank(),
         caption = element_blank())
pl_5102_patch_calepa_l2 <- pl_5102_patch_calepa_l2 + 
    plot_annotation(
        title = default_pl_title,
        caption = default_pl_caption)
ggsave(filename = here('07_slides', '2021-05-14', 'images', 
                       '5102_calepa_level2_combined.png'), 
       plot = pl_5102_patch_calepa_l2, 
       width = 10, 
       height = 6, 
       dpi = 125
       )

## 2.2. - CalEPA By Group ----

### create groupings ----
df_5102_epa_1yr <- df_5102_epa_1yr %>% 
    mutate(metrics_group = case_when(
        # legal
        employee_category == 'Legal Occupations' ~ 
            'Legal Occupations', 
        # management
        employee_category == 'Management Occupations' ~ 
            'Management Occupations', 
        # admin
        employee_category == 'Office and Administrative Support Occupations' |
            (employee_category == 'Computer and Mathematical Occupations' & 
                 sub_category == 'Unmapped Classes') ~ 
            'Administrative Occupations',
        # technical
        employee_category == 'Architecture and Engineering Occupations' |
            employee_category == 'Life, Physical, and Social Science Occupations' | 
            (employee_category == 'Computer and Mathematical Occupations' & 
                 sub_category == 'Operations Research Analysts') ~ 
            'Technical Occupations')
    )

### L1 ----
# sum_5102_groups_l1 <- df_5102_epa_1yr %>% 
#     add_count(metrics_group, # ethnicity_level1, 
#               wt = record_count,
#               name = 'group_total') %>%
#     add_count(ethnicity_level1, metrics_group,
#               wt = record_count,
#               name = 'ethnicity_group_total') %>% 
#     select(ethnicity_level1, ethnicity_group_total, group_total, metrics_group) %>%
#     distinct() %>%
#     mutate(rate = ethnicity_group_total / group_total) %>% 
#     mutate(type = 'CalEPA Workforce') %>%
#     arrange(metrics_group, ethnicity_level1) %>%
#     {.}
# 
# (pl_5102_groups_l1 <- sum_5102_groups_l1 %>% 
#     filter(!is.na(metrics_group)) %>% 
#         bind_rows(acs_data_level1)
#     ggplot() +
#     geom_bar(mapping = aes(x = ethnicity_level1, 
#                            y = rate, 
#                            fill = type), 
#              stat = 'identity', 
#              position = 'dodge') + 
#     facet_wrap(~ metrics_group) +
#     coord_flip() + 
#     scale_y_continuous(labels = percent) +
#     labs(title = 'Test', 
#          subtitle = 'Test',
#          x = 'Ethnicity Group',
#          y = 'Percent of Total', 
#          caption = 'State population data from 2019 1 yr ACS  |  workforce data from CalHR 5102') +
#     theme(legend.position = 'bottom', 
#           legend.title = element_blank()) +
#     guides(fill = guide_legend(reverse = TRUE)))

sum_5102_groups_l1 <- df_5102_epa_1yr %>% 
    filter(metrics_group == 'Legal Occupations') %>%
    fun_summary_5102_l1() %>% 
    bind_rows(acs_data_level1) %>% 
    mutate(metrics_group = 'Legal Occupations') %>%
    bind_rows(df_5102_epa_1yr %>% 
                  filter(metrics_group == 'Management Occupations') %>%
                  fun_summary_5102_l1() %>% 
                  bind_rows(acs_data_level1) %>% 
                  mutate(metrics_group = 'Management Occupations')) %>%
    bind_rows(df_5102_epa_1yr %>% 
                  filter(metrics_group == 'Administrative Occupations') %>%
                  fun_summary_5102_l1() %>% 
                  bind_rows(acs_data_level1) %>% 
                  mutate(metrics_group = 'Administrative Occupations')) %>%
    bind_rows(df_5102_epa_1yr %>% 
                  filter(metrics_group == 'Technical Occupations') %>%
                  fun_summary_5102_l1() %>% 
                  bind_rows(acs_data_level1) %>% 
                  mutate(metrics_group = 'Technical Occupations')) %>%
    mutate(type = factor(type)) %>% 
    {.}

facet_label_5102_groups_l1 <- sum_5102_groups_l1 %>% 
        filter(type == 'CalEPA Workforce') %>% 
        count(metrics_group, wt = ethnicity_total) %>% 
        mutate(facet_label = glue('{metrics_group} (CalEPA n = {n})'))


(pl_5102_groups_l1 <- sum_5102_groups_l1 %>%  
        left_join(facet_label_5102_groups_l1, 
                  by = c('metrics_group')) %>% 
        select(-metrics_group) %>% 
        rename('metrics_group' = 'facet_label') %>% 
        ggplot() +
        geom_bar(mapping = aes(x = ethnicity_level1, 
                               y = rate, 
                               fill = factor(type, levels = rev(levels(type)))), 
                 stat = 'identity', 
                 position = 'dodge') + 
        facet_wrap(~ metrics_group) +
        scale_fill_manual(values = colors_5102_state_epa) +
        scale_y_continuous(labels = label_percent(accuracy = 1L)) +
        labs(title = default_pl_title_groups,
             subtitle = default_pl_subtitle,
             x = 'Ethnicity Group',
             y = 'Percent of Total', 
             caption = default_pl_caption) +
        coord_flip() + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        guides(fill = guide_legend(reverse = TRUE)))

ggsave(filename = here('07_slides', '2021-05-14', 'images', 
                       '5102_calepa_groups_level1.png'), 
       plot = pl_5102_groups_l1, 
       width = 10, 
       height = 6, 
       dpi = 125
       )

(pl_5102_groups_l1_stack <- sum_5102_groups_l1 %>%  
        left_join(facet_label_5102_groups_l1, 
                  by = c('metrics_group')) %>% 
        select(-metrics_group) %>% 
        rename('metrics_group' = 'facet_label') %>% 
        ggplot() +
        geom_bar(mapping = aes(x = fct_rev(type), 
                               y = rate, 
                               fill = ethnicity_level1), #factor(type, levels = rev(levels(type)))), 
                 stat = 'identity') + 
        facet_wrap(~ metrics_group) +
        scale_fill_manual(values = colors_5102_2) +
        scale_y_continuous(labels = percent) +
        labs(title = default_pl_title_groups,
             subtitle = default_pl_subtitle,
             x = element_blank(),
             y = 'Percent of Total', 
             caption = default_pl_caption) +
        coord_flip() + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        guides(fill = guide_legend(reverse = TRUE)))

ggsave(filename = here('07_slides', '2021-05-14', 'images', 
                       '5102_calepa_groups_level1_stack.png'), 
       plot = pl_5102_groups_l1_stack, 
       width = 10, 
       height = 6, 
       dpi = 125
       )


### L2 ----
fun_summary_5102_groups_l2 <- function(dataset) {
    dataset %>% 
        filter(metrics_group == 'Legal Occupations') %>%
        fun_summary_5102_l2() %>% 
        bind_rows(acs_data_level2) %>% 
        mutate(metrics_group = 'Legal Occupations') %>%
        bind_rows(dataset %>% 
                      filter(metrics_group == 'Management Occupations') %>%
                      fun_summary_5102_l2() %>% 
                      bind_rows(acs_data_level2) %>% 
                      mutate(metrics_group = 'Management Occupations')) %>%
        bind_rows(dataset %>% 
                      filter(metrics_group == 'Administrative Occupations') %>%
                      fun_summary_5102_l2() %>% 
                      bind_rows(acs_data_level2) %>% 
                      mutate(metrics_group = 'Administrative Occupations')) %>%
        bind_rows(dataset %>% 
                      filter(metrics_group == 'Technical Occupations') %>%
                      fun_summary_5102_l2() %>% 
                      bind_rows(acs_data_level2) %>% 
                      mutate(metrics_group = 'Technical Occupations')) %>%
        mutate(type = factor(type)) %>% 
        {.}
}


fun_plot_5102_groups_l2 <- function(dataset, 
                                    plot_title = default_pl_title_groups, 
                                    plot_subtitle = default_pl_subtitle, 
                                    plot_caption = default_pl_caption) {
    dataset %>% 
        ggplot() +
        geom_bar(mapping = aes(x = ethnicity_level2, 
                               y = rate, 
                               fill = factor(type, levels = rev(levels(type)))), 
                 stat = 'identity', 
                 position = 'dodge') + 
        facet_wrap(~ metrics_group) +
        scale_y_continuous(labels = percent) +
        scale_fill_manual(values = colors_5102_state_epa) +
        labs(title = plot_title,
             subtitle = plot_subtitle,
             x = 'Ethnicity Group',
             y = 'Percent of Total', 
             caption = plot_caption) +
        coord_flip() + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        guides(fill = guide_legend(reverse = TRUE))
}


sum_5102_groups_l2 <- df_5102_epa_1yr %>%
    fun_summary_5102_groups_l2()

facet_label_5102_groups_l2 <- sum_5102_groups_l2 %>% 
    filter(type == 'CalEPA Workforce') %>% 
    count(metrics_group, wt = ethnicity_total) %>% 
    mutate(facet_label = glue('{metrics_group} (CalEPA n = {n})'))

(pl_5102_groups_l2 <- df_5102_epa_1yr %>% 
        fun_summary_5102_groups_l2() %>%
        left_join(facet_label_5102_groups_l2, by = c('metrics_group')) %>% 
        select(-metrics_group) %>% 
        rename('metrics_group' = 'facet_label') %>% 
        fun_plot_5102_groups_l2())

ggsave(filename = here('07_slides', '2021-05-14', 'images', 
                       '5102_calepa_groups_level2.png'), 
       plot = pl_5102_groups_l2, 
       width = 10, 
       height = 6, 
       dpi = 125
       )

        
fun_plot_5102_groups_l2_stack <- function(dataset, 
                                          plot_title = default_pl_title_groups, 
                                          plot_subtitle = default_pl_subtitle, 
                                          plot_caption = default_pl_caption) {
    dataset %>% 
        ggplot() +
        geom_bar(mapping = aes(x = fct_rev(type), 
                               y = rate, 
                               fill = ethnicity_level2), 
                 stat = 'identity') + 
        facet_wrap(~ metrics_group) +
        scale_y_continuous(labels = percent) +
        scale_fill_manual(values = colors_5102_7) +
        labs(title = plot_title,
             subtitle = plot_subtitle,
             x = element_blank(),
             y = 'Percent of Total', 
             caption = plot_caption) +
        coord_flip() + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        guides(fill = guide_legend(reverse = TRUE))
}


(pl_5102_groups_l2_stack <- df_5102_epa_1yr %>% 
        fun_summary_5102_groups_l2() %>% 
        left_join(facet_label_5102_groups_l2, by = c('metrics_group')) %>% 
        select(-metrics_group) %>% 
        rename('metrics_group' = 'facet_label') %>% 
        fun_plot_5102_groups_l2_stack())

ggsave(filename = here('07_slides', '2021-05-14', 'images', 
                       '5102_calepa_groups_level2_stack.png'), 
       plot = pl_5102_groups_l2_stack, 
       width = 10, 
       height = 6, 
       dpi = 125
       )
    

## 2.4 By BDO Plots ----
### L2 ----
summary_5102_l2_bdo <- df_5102_epa_1yr %>%
    add_count(dept,  wt = record_count,
              name = 'dept_total') %>% 
    add_count(ethnicity_level2, dept,
              wt = record_count,
              name = 'ethnicity_total') %>%
    select(ethnicity_level2, ethnicity_total, dept, dept_total) %>%
    distinct() %>%
    mutate(rate = ethnicity_total / dept_total) %>%
    mutate(type = 'CalEPA Workforce') %>%
    arrange(dept, ethnicity_level2) %>%
    mutate(type = factor(type)) %>% 
    bind_rows(acs_data_level2 %>% 
                  rename(dept = type)) %>% 
    {.}


ordering_bdo <- summary_5102_l2_bdo %>% 
    filter(ethnicity_level2 == 'White') %>% 
    arrange(rate) %>% 
    pull(dept)

(pl_5102_l2_bdo_stack <- summary_5102_l2_bdo %>% 
        # mutate(type = fct_rev(type)) %>%
        ggplot() +
        geom_bar(mapping = aes(x = fct_relevel(dept, ordering_bdo), #fct_rev(type), 
                               y = rate, 
                               fill = ethnicity_level2), # factor(type, levels = rev(levels(type)))), 
                 stat = 'identity') +
        scale_fill_manual(values = colors_5102_7) +
        scale_y_continuous(labels = percent) +
        labs(title = default_pl_title_bdo,
             x = element_blank(),
             y = 'Percent of Total', 
             caption = default_pl_caption) +
        coord_flip() + 
        # facet_wrap(~ dept) +
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        guides(fill = guide_legend(reverse = TRUE)) + 
        geom_vline(xintercept = 1.5, size = 0.5, linetype = 'dashed'))#, color = 'grey50')

ggsave(filename = here('07_slides', '2021-05-14', 'images', 
                       '5102_calepa_bdo_level2.png'), 
       plot = pl_5102_l2_bdo_stack, 
       width = 10, 
       height = 6, 
       dpi = 125
       )


### L2 - BDO & Group ----
##### enter dept name for plot vs dept name in 5102 dataset
# bdo_titles <- c("Air Resources Board" = 'Air Resources Board', 
#                 "Offc. of Env. Health Hazard Assessment" = 'Environmental Health Hazard Assessment, Office of',
#                 "Cal. Environmental Protection Agency" = 'Environmental Protection Agency',
#                 "Dept. of Pesticide Regulation" = 'Pesticide Regulation, Department of',
#                 "Dept. of Resources Recycling and Recovery" = 'Resources Recycling and Recovery, Department of',
#                 "Dept. of Toxic Substances Control" = 'Toxic Substances Control, Department of',
#                 "Water Resources Control Board" = 'Water Resources Control Board')

bdo_titles <- c("ARB" = 'Air Resources Board', 
                "OEHHA" = 'Environmental Health Hazard Assessment, Office of',
                "CalEPA (Agency)" = 'Environmental Protection Agency',
                "DPR" = 'Pesticide Regulation, Department of',
                "CalRecycle" = 'Resources Recycling and Recovery, Department of',
                "DTSC" = 'Toxic Substances Control, Department of',
                "SWRCB" = 'Water Resources Control Board')
bdo_titles_filenames <- tolower(c('ARB', 'OEHHA', 'CalEPA_Agency', 
                                  'DPR', 'CalRecycle', 'DTSC', 'SWRCB'))

# get facet labels
pl_5102_groups_l2_bdo_all_facets <- map2_df(.x = names(bdo_titles), 
                                            .y = unname(bdo_titles), 
                                            .f = ~  df_5102_epa_1yr %>% 
                                                filter(dept == .y) %>% 
                                                fun_summary_5102_groups_l2() %>% 
                                                filter(type == 'CalEPA Workforce') %>% 
                                                count(metrics_group, wt = ethnicity_total) %>% 
                                                mutate(facet_label = glue('{metrics_group} ({.x} n = {n})')) %>% 
                                                mutate(bdo = glue('{.x}'))
                                            )

#### make plots
pl_5102_groups_l2_bdo_all <- map2(.x = names(bdo_titles), 
                                  .y = unname(bdo_titles), 
                                  .f = ~ df_5102_epa_1yr %>% 
                                      filter(dept == .y) %>% 
                                      fun_summary_5102_groups_l2() %>%
                                      mutate(type = case_when(as.character(type) == 'CalEPA Workforce' ~
                                                                  glue('{.x}'),
                                                              TRUE ~ as.character(type))) %>%
                                      mutate(type = factor(type, 
                                                           levels = c(.x, 'State Population'))) %>%
                                      mutate(bdo = glue('{.x}')) %>%
                                      left_join(pl_5102_groups_l2_bdo_all_facets,
                                                by = c('metrics_group', 'bdo')) %>%
                                      select(-metrics_group) %>%
                                      rename('metrics_group' = 'facet_label') %>%
                                      distinct() %>%
                                      fun_plot_5102_groups_l2_stack(plot_title = glue('{.x} Workforce by Occupation and Ethnicity Group vs. State Population (Year {report_year})')) %>%
                                      {.}
)
## View one plot
pl_5102_groups_l2_bdo_all[[3]]

#### save plots
walk2(.x = pl_5102_groups_l2_bdo_all, 
      # .y = tolower(str_replace_all(string = names(bdo_titles), 
      #                              pattern = ' ', 
      #                              replacement = '_') %>% 
      #                  str_remove_all(pattern = '\\.|(|)')), 
      .y = bdo_titles_filenames, # keep filenames consistent regardless of changes to plot labels/titles
      .f = ~ ggsave(filename = here('07_slides', '2021-05-14', 'images', 
                                    glue('5102_bdo_group_l2_{.y}.png')), 
                    plot = .x,
                    width = 10, 
                    height = 6, 
                    dpi = 125)
)


# # plot one individual bdo (not needed if above map/walk functions are working)
# bdo_plot <- "Air Resources Board"
# bdo_plot <- 'Water Resources Control Board'
# (pl_5102_groups_l2_bdo <- df_5102_epa_1yr %>%
#     filter(dept == bdo_plot) %>%
#     fun_summary_5102_groups_l2() %>%
#     mutate(type = case_when(as.character(type) == 'CalEPA Workforce' ~
#                                 glue('{bdo_plot} Workforce'),
#                             TRUE ~ as.character(type))) %>%
#     mutate(type = factor(type, levels = c(glue('{bdo_plot} Workforce'), 'State Population'))) %>%
#     fun_plot_5102_groups_l2_stack(plot_title = glue('{bdo_plot} Workforce by Occupation and Ethnicity Group Compared to State Population (Year {report_year})'))
# )
# ggsave(filename = here('07_slides', '2021-05-14', 'images',
#                        glue('5102_bdo_group_l2_{bdo_plot}.png')),
#        plot = pl_5102_groups_l2_bdo,
#        width = 10,
#        height = 6,
#        dpi = 125
#        )


# !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
# !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
# 3 - retention / departures (by job class) ----------------------------------------------------

## get 5112 data ----
### pull data from data.ca.gov
df_5112_2020 <- read_csv('https://data.ca.gov/dataset/e620a64f-6b86-4ce0-ab4b-03d06674287b/resource/d308f328-6b5b-41c1-8bc2-a4afcfcee3d1/download/5112-inline-report_12.31.20.csv') %>% 
    clean_names()

### add additional levels for ethnicity grouping
#   ethnicity (in original dataset) is most detailed level
#   level 1 groups all BIPOC together (resulting groups are BIPOC and white)
#   level 2 groups some ethnicity sub-groupings together (Asian group and Pacific Islander group)
df_5112_2020 <- df_5112_2020 %>% 
    mutate(
        ethnicity_level1 = case_when(
            ethnicity == 'White' ~ 'White', 
            TRUE ~ 'BIPOC'
        ),
        ethnicity_level2 = case_when(
            str_detect(ethnicity, 'Pacific Islander') ~ 'Pacific Islander', # groups (all start with "Pacific Islander - "): Guamanian, Hawaiian, Other or Multiple, and Samoan
            str_detect(ethnicity, 'Asian') ~ 'Asian', # groups (all start with "Asian - "): Cambodian, Chinese, Filipino, Indian, Japanese, Korean, Laotian, Other or Multiple, Vietnamese
            TRUE ~ ethnicity)
    )

### Filtering for EPA and related BDOs 
df_5112_2020_epa <- df_5112_2020 %>% 
    filter(department == "Air Resources Board"|
               department == "Environmental Health Hazard Assessment, Office of"|
               department == "Environmental Protection Agency"|
               department == "Pesticide Regulation, Department of"|
               department == "Resources Recycling and Recovery, Department of"|
               department == "Toxic Substances Control, Department of"|
               department == "Water Resources Control Board")


## calculate... ----

job_class <- 'ENVIRONMENTAL SCIENTIST'

job_class_flux <- df_5112_2020_epa %>% 
    filter(class_title == job_class) %>% 
    group_by(ethnicity_level2, hire_type) %>% 
    summarize(count = n()) %>% 
    ungroup() %>% 
    mutate(hire_type = case_when(
        hire_type == 'Intake: Outside Hires' ~ 
            'intake_outside',
        hire_type == 'Intake: Hires from other State Agencies' ~ 
            'intake_other_agencies',
        hire_type == 'Advancements' ~ 
            'intake_advancements'
    ))
    # add_count(ethnicity, name = 'ethnicity_total') %>% 
    # add_count(ethnicity, hire_type, name = 'ethnicity_type_total')

job_class_flux <- job_class_flux %>% 
    pivot_wider(names_from = hire_type, 
                values_from = count)

job_class_delta <- df_5102_epa %>% 
    filter(class_title == job_class) %>% 
    filter(as_of_date == '2020-12-31' | as_of_date == '2019-12-31') %>% 
    mutate(year = glue('yr_{year(as_of_date)}')) %>% 
    group_by(year, ethnicity_level2) %>% 
    summarize(total = sum(record_count)) %>% 
    ungroup() %>% 
    pivot_wider(names_from = year, values_from = total) %>% 
    mutate(change = yr_2020 - yr_2019) %>% 
    {.}
    
job_class_calc <- job_class_delta %>% 
    left_join(job_class_flux, by = 'ethnicity_level2') %>% 
    mutate_if(is.numeric, ~replace_na(., 0)) %>% 
    mutate(intake_total = intake_outside + 
               intake_other_agencies + 
               intake_advancements) %>% 
    mutate(departure = intake_total - change)


## timeline ----
job_class_timeline <- df_5102_epa %>% 
    filter(class_title == job_class) 

```
