---
title: "5102 Data Analysis by Agency / Department"
author: "CalEPA Racial Equity Team"
date: "*Updated: `r format(Sys.Date())`*"
output:
  html_document:
    code_folding: show
    toc: TRUE
    toc_float: FALSE
    toc_depth: 3
runtime: shiny
assets:
  css:
    - "http://fonts.googleapis.com/css?family=Raleway:300"
    - "http://fonts.googleapis.com/css?family=Oxygen"
---

<style>
body{
  font-family: 'Oxygen', sans-serif;
  font-size: 16px;
  line-height: 24px;
}

h1,h2,h3,h4 {
  font-family: 'Raleway', sans-serif;
}

.container { width: 1250px; }
h3 {
  background-color: #D4DAEC;
  text-indent: 50px; 
}
h4 {
  text-indent: 75px;
  margin-top: 35px;
  margin-bottom: 5px;
}
</style>

# 5102 Data Analysis by Agency / Department

## Introduction
This code can be modified and applied to any department(s) in the California
state system listed in California Department of Human Resource's (CalHR) 5102 data 
set. This code was written by California Environmental Protection Agency's (CalEPA) 
Racial Equity Team (CRET) Data Subteam in Spring 2021 to provide helpful 
racial demographics visualizations for the workforce of the Agency and its Boards, 
Departments, and Office (BDOs). The effort came out of a desire to supplement the
Agency's Race Forward survey results with workforce demographic data. The Data 
Subteam intends on maintaining and building off this and other resources and datasets.
The vision of the Workforce Equity Subteam's work is to create a workforce in which 
one's work and/or career outcomes are not predicted by one's race. The Data Subteam's 
efforts support this vision. This vision is informed by more detailed goals and objectives 
related to work/career satisfaction, promotion and retention, compensation, 
inclusion, mental and other health safety, and more. 
CRET's model of change - normalize, organize and operationalize – is 
aligned with over 200 government organizations part of a nationwide network to 
transform governments to deliver more racial equity. We anticipate that these
demographic metrics and resulting performance measures will evolve over time and 
will be informed by future data and analyses of surveys, workforce demographics, 
and research.
Finally, the Data Subteam acknowledges that racial equity work is complex and 
cannot be pared down to simple numbers in bins. This code and resulting 
visualizations seeks to be the first step in critically examining an agency's 
racial demographics, and may be more helpful in developing the right questions 
that need to be asked, rather than immediate answers and/or numeric targets.
If you have any questions about this code, please email Leah Jones at <leah.jones@waterboards.ca.gov>.

## CA Workforce Data Resources

-   CalHR data page:
    <https://www.calhr.ca.gov/pages/statewide-reports.aspx>

-   Workforce data on CA open data portal:
    <https://data.ca.gov/dataset/calhr-civil-rights-data-for-gare-capital-cohort-2019>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE) # set "echo = FALSE" to prevent all code from displaying in the output
```


```{r Libraries, include=TRUE, message=FALSE}
# If you need to install the packages first, use the following code:
# install.packages("readxl")
# install.packages("tidyverse")
# install.packages("janitor")
# install.packages("here")
# install.packages("glue")
# install.packages("lubridate")
# install.packages("scales")
# install.packages("tidycensus")
# install.packages("sf")
# install.packages("patchwork")
# install.packages("knitr")
# install.packages("kableExtra")
# install.packages("DT")

# load packages ----
library(readxl)
library(tidyverse)
library(janitor)
library(here)
library(glue)
library(lubridate)
library(scales)
library(tidycensus)
library(sf)
library(patchwork)
library(knitr)      # generate html tables for this report
library(kableExtra) # better formatting of tables
library(DT)         # creates sortable tables
```

## Enter Parameters

Enter the year of the most recently published 5102 report and the most
recently published ACS data, and if needed edit the URL for the 5102
data set on the California Open Data Portal. Also, edit default plot
titles and captions as needed (CalEPA and its 6 Boards, Departments, 
and Offices [BDOs] are the examples used here).

```{r}
# enter parameters (for current reporting year) -----------------------------
report_year <- 2020
acs_data_year <- 2019

# enter agency name
agency_name <- 'CalEPA'

url_data_all_yrs <- 'https://data.ca.gov/dataset/e620a64f-6b86-4ce0-ab4b-03d06674287b/resource/aba87ad9-f6b0-4a7e-a45e-d1452417eb7f/download/calhr_5102_statewide_2011-2020.csv'

default_pl_title <- glue('{agency_name} Workforce Compared to State Population (Year {report_year})')
default_pl_title_groups <- glue('{agency_name} Workforce by Occupation and Ethnicity Group Compared to State Population (Year {report_year})')
default_pl_title_bdo <- glue('{agency_name} Workforce by BDO Compared to State Population (Year {report_year})')

default_pl_subtitle <- element_blank()
default_pl_caption <- glue('Data sources: state population data from {acs_data_year} 1 yr American Community Survey  |  workforce data from {report_year} CalHR 5102 Report')
```

## Load & Transform Data

### 5102 Data

CalHR's biannual 5102 report "captures the race/ethnicity, gender, disability, 
and veteran status by occupational group and classification for all state employees."
Data source: [CalHR's Statewide 5102 report](https://www.calhr.ca.gov/pages/statewide-reports.aspx). 
A cleaned and compiled version of that data is available on the California Open Data Portal at: <https://data.ca.gov/dataset/calhr-civil-rights-data-for-gare-capital-cohort-2019>

#### Ethnicity Groupings

The ethnicity groupings were thought of in two levels. Level one is "white" versus
"BIPOC", where BIPOC stands for black, indigenous, and people of color, and includes 
all ethnicity values other than white. Level two ethnicity groups some sub-groupings,
in this case all Asian identities. We recognize that all ethnicities, including
sub-groups of Asian ethnicities, experience discrimination differently. However,
for the purposes of creating useful visualizations with a relatively small dataset, 
we chose to group the identity variables in this way. Feel free to explore the
data and re-group for your purposes as you see fit.

```{r}
# 1 - load / transform data ---------------------------------------------------------------

## 5102 ----
df_5102 <- read_csv(url_data_all_yrs) %>%
    clean_names() 

# add additional levels for ethnicity grouping
#   identity_variable (in original dataset) is most detailed level
#   level 1 groups all BIPOC together (resulting groups are BIPOC and white)
#   level 2 groups some identity_variable sub-groupings together (Asian group and Pacific Islander group)
df_5102 <- df_5102 %>% 
    mutate(
        ethnicity_level1 = case_when(
            identity_variable == 'White' ~ 'White', 
            TRUE ~ 'BIPOC'),
        ethnicity_level2 = case_when(
            str_detect(identity_variable, 'Pacific Islander') ~ 'Pacific Islander', # groups (all start with "Pacific Islander - "): Guamanian, Hawaiian, Other or Multiple, and Samoan
            str_detect(identity_variable, 'Asian') ~ 'Asian', # groups (all start with "Asian - "): Cambodian, Chinese, Filipino, Indian, Japanese, Korean, Laotian, Other or Multiple, Vietnamese
            TRUE ~ identity_variable)
    )

# add a year column
df_5102 <- df_5102 %>% 
    mutate(reporting_year = year(as_of_date))
```

#### Filter for Selected Agency

Search for the name of your selected agency from the box below, then copy and 
paste into the subsequent code chunk.

List of agency names in the 5102 data set:

```{r}
# list of all departments
depts_list <- df_5102 %>% 
    distinct(dept) %>% 
    arrange(dept) %>% 
    rename('Department Name' = dept)


depts_list %>%
    kbl() %>%
    kable_paper() %>%
    scroll_box(width = "500px", height = "200px") %>%
    {.}
```


TESTING SHINY
# ```{r}
# ui <- fluidPage(
#   selectInput('Department Name', label='Select Department', choices=names(dept), selected=1),
#   plotOutput('distPlot')
# )
# 
# server <- function(input, output, session) {
#   output$distPlot <- renderPlot({
#   plot(dept[[input$distName]]$fitmodel) # view plot fits
#     title(input$distName, outer=T)
# })
# }
# 
# shinyApp(ui, server, options=list(height=500))
# )
# ```

Enter departments to be included in the analysis, and modify the variable 
names as you see fit:

```{r}
# Filtering for EPA and related BDOs 
df_5102_dept <- df_5102 %>% 
    filter(dept == "Air Resources Board"|
               dept == "Environmental Health Hazard Assessment, Office of"|
               dept == "Environmental Protection Agency"|
               dept == "Pesticide Regulation, Department of"|
               dept == "Resources Recycling and Recovery, Department of"|
               dept == "Toxic Substances Control, Department of"|
               dept == "Water Resources Control Board")

# filter for data from the report year (selected above)
df_5102_dept_1yr <- df_5102_dept %>% 
    filter(reporting_year == report_year)

```

### Census Data

For this report, we use data from the U.S. Census Bureau's 1-year American Community Survey (ACS) dataset, which provides a statewide estimate of population and demographic data for a given year. The selected year (entered above) should match the year of the workforce data being analyzed, or if the ACS data for the desired year is not yet available the previous year's 1-year ACS dataset can be used. To access the ACS data we use the tidycensus R package, which provides a user friendly interface to to the Census Bureau's API via R. 

The ACS dataset provides information about a wide variety of topics, with different data types provided in separate tables, and fields within those tables defined by codes. We used ACS table B03002 to obtain estimates of statewide population by racial/ethnic group, because that table includes information about hispanic/latino origin (note that the primary ACS table for race/ethnicity information is table B02001, but that table does not contain information about hispanic/latino origin, as hispanic/latino orginin considered to be a characteristic that is tracked independent of race). 

The ethnicity groupings provided in ACS table B03002 are roughly analogous to the "level 2" ethnicity groupings described above for the state workforce data. To make the census data consistent with the level 2 groupings state workforce data, the only transformation applied to the census data is grouping the "Other" and "Multiple" category into one 'Other or Multiple Race' category. We also create an ACS dataset consistent with the level 1 ethnicity groupings described above by combining all of the racial/ethnic categories except "White" into one "BIOPOC" category.

For more information about the tidycensus R package or the various types of data available from the U.S. Census Bureau, see the resources listed below.

#### Census Data Resources

-   tidycensus R package:
    <https://walker-data.com/tidycensus/articles/basic-usage.html>

-   general census workshop w/ R:
    <https://jennhuck.github.io/workshops/tidycensus.html>

-   ACS Table B02001 data (race w/o hispanic/latino data):
    <https://censusreporter.org/data/table/?table=B02001&geo_ids=04000US06>

-   ACS Table B03002 data (race with hispanic/latino data):
    <https://censusreporter.org/data/table/?table=B03002&geo_ids=04000US06>

-   info about race in census data:
    <https://censusreporter.org/topics/race-hispanic/>

-   other general info about concepts and tables in census data:
    <https://censusreporter.org/topics/>

-   check/validation of the numbers for CA:
    <https://www.ppic.org/publication/californias-population/>

#### Download Census Data

To download the census data, you need to first request a key here: <https://api.census.gov/data/key_signup.html/>
Then, enter census_api_key("*your_key*", install = TRUE) in your console.


```{r}
## census data (ACS 1 yr) ----

### get ACS data
acs_data_raw <- get_acs(geography = 'state', 
                        variables = c(# total_state_pop = 'B02001_001',
                            'Hispanic or Latino' = 'B03002_012', # Total Hispanic or Latino
                            'White' = 'B03002_003', # White (Not Hispanic or Latino)
                            'Black or African American' = 'B03002_004', # Black or African American (Not Hispanic or Latino)
                            'Native American or Alaska Native' = 'B03002_005', # American Indian and Alaska Native (Not Hispanic or Latino)
                            'Asian' = 'B03002_006', # Asian (Not Hispanic or Latino)
                            'Pacific Islander' = 'B03002_007', # Native Hawaiian and Other Pacific Islander (Not Hispanic or Latino)
                            'Other' = 'B03002_008', # Some other race (Not Hispanic or Latino)
                            'Multiple' = 'B03002_009'# Two or more races (Not Hispanic or Latino)
                        ),
                        summary_var = c(total_state_pop = 'B02001_001'), 
                        survey = 'acs1', # use 'acs1' or 'acs5' to get 1 or 5 year acs
                        state = 'CA', 
                        geometry = FALSE, # set to TRUE to get as geospatial data
                        year = acs_data_year)

# clean / reformat the acs data (the data will be formatted to be 
# consistent with the "level 2" ethnicity groupings in the workforce 
# data that are created above)
acs_data_level2 <- acs_data_raw %>% 
    clean_names() %>% 
    rename(total_state_pop = summary_est, 
           total_state_pop_moe = summary_moe,
           location_name = name,
           ethnicity_level2 = variable)

    ## check (should be TRUE) - make sure sum of populations by group == total state population
    # sum(acs_data_level2$estimate) == acs_data_level2$total_state_pop[1]

### group the 'Other' and 'Multiple' rows into one 'Other or Multiple Race' row
acs_data_level2 <- acs_data_level2 %>% 
    mutate(ethnicity_level2 = 
               case_when(ethnicity_level2 == 'Other' | 
                             ethnicity_level2 == 'Multiple' ~ 
                             'Other or Multiple Race',
                         TRUE ~ ethnicity_level2)) %>% 
    group_by(geoid, location_name, ethnicity_level2, total_state_pop) %>% 
    summarize(estimate = sum(estimate)) %>%
    ungroup() %>% 
    {.}
    
    ## check (should be TRUE) - make sure sum of populations by group == total state population
    # sum(acs_data_level2$estimate) == acs_data_level2$total_state_pop[1]
    
### add a column with each ethnicity's % of total state population
acs_data_level2 <- acs_data_level2 %>% 
    mutate(pop_pct = estimate / total_state_pop) %>% 
    {.}

### create a dataset grouped at level 1 - all BIPOC together (resulting groups are BIPOC and white)
acs_data_level1 <- acs_data_level2 %>%  
    mutate(
        ethnicity_level1 = case_when(
            ethnicity_level2 == 'White' ~ 'White', 
            TRUE ~ 'BIPOC')
    ) %>% 
    group_by(geoid, location_name, ethnicity_level1, total_state_pop) %>% 
    summarize(estimate = sum(estimate)) %>% 
    ungroup() %>% 
    mutate(pop_pct = estimate / total_state_pop) %>% # update the pop_pct to reflect level 1 numbers
    {.}
    
    ## check 
    # sum(acs_data_level1$estimate) == acs_data_level1$total_state_pop[1]
```

## Plots

### Enter Plotting Parameters

Here you can modify the plot color palette and other parameters:

```{r}
# !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
# !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
# 2 - 5102 summaries (vs CA pop) -----------------------------------------------
## define color palette for plots
### state pop vs dept workforce plots
colors_5102_state_dept <-c('gold', 'darkgreen') # state pop / dept workforce
### colors for ethnicities (in stacked bar plots) - trying to keep 'white' the same for both levels of plots for consistency/comparison
brewer_colors <- rev(RColorBrewer::brewer.pal(n = 7, name = 'Set2'))
colors_5102_2 <- brewer_colors[6:7] # BIPOC / White
colors_5102_7 <- brewer_colors # detailed ethnic groups: Asian ... White
    

```


### Format Data & Create Plotting Functions

No changes necessary here, unless you would like to update variable and/or title 
names:

```{r}

## format census data for plot (all of CA) ----
acs_data_level1 <- acs_data_level1 %>%
    mutate(type = factor('State Population')) %>%
    rename(ethnicity_total = estimate,
           rate = pop_pct) %>%
    
    {.}

acs_data_level2 <- acs_data_level2 %>%
    mutate(type = factor('State Population')) %>%
    rename(ethnicity_total = estimate,
           rate = pop_pct) %>%
    {.}

## 5102 summary functions
fun_summary_5102_l1 <- function(dataset) {
    dataset %>%
        add_count(ethnicity_level1,
                  wt = record_count,
                  name = 'ethnicity_total') %>%
        select(ethnicity_level1, ethnicity_total) %>%
        distinct() %>%
        mutate(rate = ethnicity_total / sum(ethnicity_total)) %>%
        mutate(type = 'Department Workforce') %>%
        arrange(ethnicity_level1) %>%
        mutate(type = factor(type)) %>% 
        {.}
}

fun_summary_5102_l2 <- function(dataset) {
    dataset %>%
        add_count(ethnicity_level2,
                  wt = record_count,
                  name = 'ethnicity_total') %>%
        select(ethnicity_level2, ethnicity_total) %>%
        distinct() %>%
        mutate(rate = ethnicity_total / sum(ethnicity_total)) %>%
        mutate(type = 'Department Workforce') %>%
        arrange(ethnicity_level2) %>%
        mutate(type = factor(type)) %>% 
        {.}
}

## plot function
fun_plot_5102_l1 <- function(dataset, plot_title, plot_caption) {
    dataset %>% 
        ggplot() +
        geom_bar(mapping = aes(x = ethnicity_level1, 
                               y = rate, 
                               fill = factor(type, levels = rev(levels(type)))), 
                 stat = 'identity', 
                 position = 'dodge') +
        scale_fill_manual(values = colors_5102_state_dept) +
        scale_y_continuous(labels = label_percent(accuracy = 1L)) +
        labs(title = plot_title,
             x = 'Ethnicity Group',
             y = 'Percent of Total', 
             caption = plot_caption) +
        coord_flip() + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        guides(fill = guide_legend(reverse = TRUE))
}

fun_plot_5102_l1_stack <- function(dataset, plot_title, plot_caption) {
    dataset %>% 
        mutate(type = fct_rev(type)) %>% 
        ggplot() +
        geom_bar(mapping = aes(x = type, # ethnicity_level1, 
                               y = rate, 
                               fill = ethnicity_level1), # factor(type, levels = rev(levels(type)))), 
                 stat = 'identity') +
        scale_fill_manual(values = colors_5102_2) +
        scale_y_continuous(labels = label_percent(accuracy = 1L)) +
        labs(title = plot_title,
             x = element_blank(),
             y = 'Percent of Total', 
             caption = plot_caption) + 
        coord_flip() + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        guides(fill = guide_legend(reverse = TRUE))
}


fun_plot_5102_l2 <- function(dataset, plot_title, plot_caption) {
    dataset %>% 
        ggplot() +
        geom_bar(mapping = aes(x = ethnicity_level2, 
                               y = rate, 
                               fill = factor(type, levels = rev(levels(type)))), 
                 stat = 'identity', 
                 position = 'dodge') +
        scale_fill_manual(values = colors_5102_state_dept) +
        scale_y_continuous(labels = label_percent(accuracy = 1L)) +
        labs(title = plot_title,
             x = 'Ethnicity Group',
             y = 'Percent of Total', 
             caption = plot_caption) +
        coord_flip() + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        guides(fill = guide_legend(reverse = TRUE))
}

fun_plot_5102_l2_stack <- function(dataset, plot_title, plot_caption) {
    dataset %>% 
        mutate(type = fct_rev(type)) %>%
        ggplot() +
        geom_bar(mapping = aes(x = type, 
                               y = rate, 
                               fill = ethnicity_level2), # factor(type, levels = rev(levels(type)))), 
                 stat = 'identity') +
        scale_fill_manual(values = colors_5102_7) +
        scale_y_continuous(labels = label_percent(accuracy = 1L)) +
        labs(title = plot_title,
             x = element_blank(),
             y = 'Percent of Total', 
             caption = plot_caption) +
        coord_flip() + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        guides(fill = guide_legend(reverse = TRUE))
}
```


### Department Summary Plots

*To save the graphs as png image files, remove the comment symbols (#) from the* 
*ggsave lines.*


#### Level 1 Ethnicity

```{r}
#### Level 1 ---- 
(pl_5102_dept_l1 <- df_5102_dept_1yr %>% 
     fun_summary_5102_l1() %>% 
     bind_rows(acs_data_level1) %>% 
     fun_plot_5102_l1(plot_title = default_pl_title, 
                      plot_caption = default_pl_caption))

(pl_5102_dept_l1_stack <- df_5102_dept_1yr %>% 
        fun_summary_5102_l1() %>% 
        bind_rows(acs_data_level1) %>% 
        fun_plot_5102_l1_stack(plot_title = default_pl_title, 
                               plot_caption = default_pl_caption))

# ggsave(filename = here('07_slides', '2021-05-14', 'images', 
#                        '5102_dept_level1_stack.png'), 
#        plot = pl_5102_dept_l1_stack, 
#        width = 10, 
#        height = 6, 
#        dpi = 125
#        )

# # patchwork (combine plots) - Level 1
# pl_5512_patch_dept_l1 <- pl_5102_dept_l1_stack / pl_5102_dept_l1
```

#### Level 2 Ethnicity

```{r}
### Level 2 ----
(pl_5102_dept_l2 <- df_5102_dept_1yr %>% 
     fun_summary_5102_l2() %>% 
     bind_rows(acs_data_level2) %>% 
     fun_plot_5102_l2(plot_title = default_pl_title, 
                      plot_caption = default_pl_caption))
 
# ggsave(filename = here('07_slides', '2021-05-14', 'images', 
#                        '5102_dept_level2.png'), 
#        plot = pl_5102_dept_l2, 
#        width = 10, 
#        height = 6, 
#        dpi = 125
#        )

(pl_5102_dept_l2_stack <- df_5102_dept_1yr %>% 
        fun_summary_5102_l2() %>% 
        bind_rows(acs_data_level2) %>% 
        fun_plot_5102_l2_stack(plot_title = default_pl_title, 
                               plot_caption = default_pl_caption))

# ggsave(filename = here('07_slides', '2021-05-14', 'images', 
#                        '5102_dept_level2_stack.png'), 
#        plot = pl_5102_dept_l2_stack, 
#        width = 10, 
#        height = 6, 
#        dpi = 125
#        )

# patchwork (combine plots) - Level 1 and Level 2 stacked plots
pl_5102_patch_dept_stack_combined <- pl_5102_dept_l1_stack / pl_5102_dept_l2_stack
pl_5102_patch_dept_stack_combined[[1]] <- pl_5102_patch_dept_stack_combined[[1]] + 
    labs(title = element_blank(),
         subtitle = element_blank(),
         caption = element_blank())
pl_5102_patch_dept_stack_combined[[2]] <- pl_5102_patch_dept_stack_combined[[2]] + 
    labs(title = element_blank(),
         subtitle = element_blank(),
         caption = element_blank())
pl_5102_patch_dept_stack_combined <- pl_5102_patch_dept_stack_combined + 
    plot_layout(heights = c(1.7, 2)) +
    plot_annotation(
        title = default_pl_title,
        caption = default_pl_caption)
# ggsave(filename = here('07_slides', '2021-05-14', 'images', 
#                        '5102_dept_level1_2_combined_stack.png'), 
#        plot = pl_5102_patch_dept_stack_combined, 
#        width = 10, 
#        height = 6, 
#        dpi = 125
#        )

# patchwork (combine plots) - both L2 plots
pl_5102_patch_dept_l2 <- pl_5102_dept_l2_stack / pl_5102_dept_l2
pl_5102_patch_dept_l2[[1]] <- pl_5102_patch_dept_l2[[1]] + 
    labs(title = element_blank(),
         subtitle = element_blank(),
         caption = element_blank())
pl_5102_patch_dept_l2[[2]] <- pl_5102_patch_dept_l2[[2]] + 
    labs(title = element_blank(),
         subtitle = element_blank(),
         caption = element_blank())
pl_5102_patch_dept_l2 <- pl_5102_patch_dept_l2 + 
    plot_annotation(
        title = default_pl_title,
        caption = default_pl_caption)
# ggsave(filename = here('07_slides', '2021-05-14', 'images', 
#                        '5102_dept_level2_combined.png'), 
#        plot = pl_5102_patch_dept_l2, 
#        width = 10, 
#        height = 6, 
#        dpi = 125
#        )
```

### Department by Employee Class

To compare across employee categories, use the code chunks below to identify your 
desired groupings and modify the categories/classes as you see fit.

```{r}
## 2.2. - Department By Group ----

# To view list of all categories
categories_list <- df_5102 %>% 
  distinct(employee_category,sub_category, class_title) %>% 
  arrange(employee_category) %>% 
  rename('Employee Category' = employee_category) %>%
  rename('Sub-Category' = sub_category) %>% 
  rename('Class Title' = class_title)


categories_list %>% 
    kbl() %>% 
    kable_paper() %>%
    scroll_box(width = "500px", height = "200px") %>%
    {.}
```

Create your employee class groupings in the code below. Investigate the table 
above and code below to determine how you would like to group your classes.
In our example, we grouped by Legal, Management, Technical, Administrative,
and Technical, because we wanted to see the differences in demographics between
these groups. We chose these groupings to investigate assumptions about diversity
in these varying categories.

```{r}
## you can also use the code below to view your table in RStudio, to examine
## what classes and are in your department:
#View(df_5102_dept_1yr)

### create groupings ----
df_5102_dept_1yr <- df_5102_dept_1yr %>% 
    mutate(metrics_group = case_when(
        # legal
        employee_category == 'Legal Occupations' ~ 
            'Legal Occupations', 
        # management
        employee_category == 'Management Occupations' ~ 
            'Management Occupations', 
        # admin
        employee_category == 'Office and Administrative Support Occupations' |
            (employee_category == 'Computer and Mathematical Occupations' & 
                 sub_category == 'Unmapped Classes') ~ 
            'Administrative Occupations',
        # technical
        employee_category == 'Architecture and Engineering Occupations' |
            employee_category == 'Life, Physical, and Social Science Occupations' | 
            (employee_category == 'Computer and Mathematical Occupations' & 
                 sub_category == 'Operations Research Analysts') ~ 
            'Technical Occupations')
    )
```


#### Level 1 Ethnicity

```{r}
# Level 1 ethnicity
sum_5102_groups_l1 <- df_5102_dept_1yr %>% 
    filter(metrics_group == 'Legal Occupations') %>%
    fun_summary_5102_l1() %>% 
    bind_rows(acs_data_level1) %>% 
    mutate(metrics_group = 'Legal Occupations') %>%
    bind_rows(df_5102_dept_1yr %>% 
                  filter(metrics_group == 'Management Occupations') %>%
                  fun_summary_5102_l1() %>% 
                  bind_rows(acs_data_level1) %>% 
                  mutate(metrics_group = 'Management Occupations')) %>%
    bind_rows(df_5102_dept_1yr %>% 
                  filter(metrics_group == 'Administrative Occupations') %>%
                  fun_summary_5102_l1() %>% 
                  bind_rows(acs_data_level1) %>% 
                  mutate(metrics_group = 'Administrative Occupations')) %>%
    bind_rows(df_5102_dept_1yr %>% 
                  filter(metrics_group == 'Technical Occupations') %>%
                  fun_summary_5102_l1() %>% 
                  bind_rows(acs_data_level1) %>% 
                  mutate(metrics_group = 'Technical Occupations')) %>%
    mutate(type = factor(type)) %>% 
    {.}

facet_label_5102_groups_l1 <- sum_5102_groups_l1 %>% 
    filter(type == 'Department Workforce') %>% 
    count(metrics_group, wt = ethnicity_total) %>% 
    mutate(facet_label = glue('{metrics_group} (Department n = {n})'))


(pl_5102_groups_l1 <- sum_5102_groups_l1 %>%  
        left_join(facet_label_5102_groups_l1, 
                  by = c('metrics_group')) %>% 
        select(-metrics_group) %>% 
        rename('metrics_group' = 'facet_label') %>% 
        ggplot() +
        geom_bar(mapping = aes(x = ethnicity_level1, 
                               y = rate, 
                               fill = factor(type, levels = rev(levels(type)))), 
                 stat = 'identity', 
                 position = 'dodge') + 
        facet_wrap(~ metrics_group) +
        scale_fill_manual(values = colors_5102_state_dept) +
        scale_y_continuous(labels = label_percent(accuracy = 1L)) +
        labs(title = default_pl_title_groups,
             subtitle = default_pl_subtitle,
             x = 'Ethnicity Group',
             y = 'Percent of Total', 
             caption = default_pl_caption) +
        coord_flip() + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        guides(fill = guide_legend(reverse = TRUE)))

# ggsave(filename = here('07_slides', '2021-05-14', 'images', 
#                        '5102_dept_groups_level1.png'), 
#        plot = pl_5102_groups_l1, 
#        width = 10, 
#        height = 6, 
#        dpi = 125
#        )

(pl_5102_groups_l1_stack <- sum_5102_groups_l1 %>%  
        left_join(facet_label_5102_groups_l1, 
                  by = c('metrics_group')) %>% 
        select(-metrics_group) %>% 
        rename('metrics_group' = 'facet_label') %>% 
        ggplot() +
        geom_bar(mapping = aes(x = fct_rev(type), 
                               y = rate, 
                               fill = ethnicity_level1), #factor(type, levels = rev(levels(type)))), 
                 stat = 'identity') + 
        facet_wrap(~ metrics_group) +
        scale_fill_manual(values = colors_5102_2) +
        scale_y_continuous(labels = percent) +
        labs(title = default_pl_title_groups,
             subtitle = default_pl_subtitle,
             x = element_blank(),
             y = 'Percent of Total', 
             caption = default_pl_caption) +
        coord_flip() + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        guides(fill = guide_legend(reverse = TRUE)))

# ggsave(filename = here('07_slides', '2021-05-14', 'images', 
#                        '5102_dept_groups_level1_stack.png'), 
#        plot = pl_5102_groups_l1_stack, 
#        width = 10, 
#        height = 6, 
#        dpi = 125
#        )
```


#### Level 2 Ethnicity

```{r}
### Level 2 ----
fun_summary_5102_groups_l2 <- function(dataset) {
    dataset %>% 
        filter(metrics_group == 'Legal Occupations') %>%
        fun_summary_5102_l2() %>% 
        bind_rows(acs_data_level2) %>% 
        mutate(metrics_group = 'Legal Occupations') %>%
        bind_rows(dataset %>% 
                      filter(metrics_group == 'Management Occupations') %>%
                      fun_summary_5102_l2() %>% 
                      bind_rows(acs_data_level2) %>% 
                      mutate(metrics_group = 'Management Occupations')) %>%
        bind_rows(dataset %>% 
                      filter(metrics_group == 'Administrative Occupations') %>%
                      fun_summary_5102_l2() %>% 
                      bind_rows(acs_data_level2) %>% 
                      mutate(metrics_group = 'Administrative Occupations')) %>%
        bind_rows(dataset %>% 
                      filter(metrics_group == 'Technical Occupations') %>%
                      fun_summary_5102_l2() %>% 
                      bind_rows(acs_data_level2) %>% 
                      mutate(metrics_group = 'Technical Occupations')) %>%
        mutate(type = factor(type)) %>% 
        {.}
}


fun_plot_5102_groups_l2 <- function(dataset, 
                                    plot_title = default_pl_title_groups, 
                                    plot_subtitle = default_pl_subtitle, 
                                    plot_caption = default_pl_caption) {
    dataset %>% 
        ggplot() +
        geom_bar(mapping = aes(x = ethnicity_level2, 
                               y = rate, 
                               fill = factor(type, levels = rev(levels(type)))), 
                 stat = 'identity', 
                 position = 'dodge') + 
        facet_wrap(~ metrics_group) +
        scale_y_continuous(labels = percent) +
        scale_fill_manual(values = colors_5102_state_dept) +
        labs(title = plot_title,
             subtitle = plot_subtitle,
             x = 'Ethnicity Group',
             y = 'Percent of Total', 
             caption = plot_caption) +
        coord_flip() + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        guides(fill = guide_legend(reverse = TRUE))
}


sum_5102_groups_l2 <- df_5102_dept_1yr %>%
    fun_summary_5102_groups_l2()

facet_label_5102_groups_l2 <- sum_5102_groups_l2 %>% 
    filter(type == 'Department Workforce') %>% 
    count(metrics_group, wt = ethnicity_total) %>% 
    mutate(facet_label = glue('{metrics_group} (Department n = {n})'))

(pl_5102_groups_l2 <- df_5102_dept_1yr %>% 
        fun_summary_5102_groups_l2() %>%
        left_join(facet_label_5102_groups_l2, by = c('metrics_group')) %>% 
        select(-metrics_group) %>% 
        rename('metrics_group' = 'facet_label') %>% 
        fun_plot_5102_groups_l2())

# ggsave(filename = here('07_slides', '2021-05-14', 'images', 
#                        '5102_dept_groups_level2.png'), 
#        plot = pl_5102_groups_l2, 
#        width = 10, 
#        height = 6, 
#        dpi = 125
#        )

        
fun_plot_5102_groups_l2_stack <- function(dataset, 
                                          plot_title = default_pl_title_groups, 
                                          plot_subtitle = default_pl_subtitle, 
                                          plot_caption = default_pl_caption) {
    dataset %>% 
        ggplot() +
        geom_bar(mapping = aes(x = fct_rev(type), 
                               y = rate, 
                               fill = ethnicity_level2), 
                 stat = 'identity') + 
        facet_wrap(~ metrics_group) +
        scale_y_continuous(labels = percent) +
        scale_fill_manual(values = colors_5102_7) +
        labs(title = plot_title,
             subtitle = plot_subtitle,
             x = element_blank(),
             y = 'Percent of Total', 
             caption = plot_caption) +
        coord_flip() + 
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        guides(fill = guide_legend(reverse = TRUE))
}


(pl_5102_groups_l2_stack <- df_5102_dept_1yr %>% 
        fun_summary_5102_groups_l2() %>% 
        left_join(facet_label_5102_groups_l2, by = c('metrics_group')) %>% 
        select(-metrics_group) %>% 
        rename('metrics_group' = 'facet_label') %>% 
        fun_plot_5102_groups_l2_stack())

# ggsave(filename = here('07_slides', '2021-05-14', 'images', 
#                        '5102_dept_groups_level2_stack.png'), 
#        plot = pl_5102_groups_l2_stack, 
#        width = 10, 
#        height = 6, 
#        dpi = 125
#        )
```

### Department/BDO Comparison

To include different levels of your agency, use the code below to compare. You could also compare departments across the state system.

#### Level 2 Ethnicity

```{r}
## 2.4 By BDO Plots ----
### Level 2 ethnicity ----
summary_5102_l2_bdo <- df_5102_dept_1yr %>%
    add_count(dept,  wt = record_count,
              name = 'dept_total') %>% 
    add_count(ethnicity_level2, dept,
              wt = record_count,
              name = 'ethnicity_total') %>%
    select(ethnicity_level2, ethnicity_total, dept, dept_total) %>%
    distinct() %>%
    mutate(rate = ethnicity_total / dept_total) %>%
    mutate(type = 'Department Workforce') %>%
    arrange(dept, ethnicity_level2) %>%
    mutate(type = factor(type)) %>% 
    bind_rows(acs_data_level2 %>% 
                  rename(dept = type)) %>% 
    {.}


ordering_bdo <- summary_5102_l2_bdo %>% 
    filter(ethnicity_level2 == 'White') %>% 
    arrange(rate) %>% 
    pull(dept)

(pl_5102_l2_bdo_stack <- summary_5102_l2_bdo %>% 
        # mutate(type = fct_rev(type)) %>%
        ggplot() +
        geom_bar(mapping = aes(x = fct_relevel(dept, ordering_bdo), #fct_rev(type), 
                               y = rate, 
                               fill = ethnicity_level2), # factor(type, levels = rev(levels(type)))), 
                 stat = 'identity') +
        scale_fill_manual(values = colors_5102_7) +
        scale_y_continuous(labels = percent) +
        labs(title = default_pl_title_bdo,
             x = element_blank(),
             y = 'Percent of Total', 
             caption = default_pl_caption) +
        coord_flip() + 
        # facet_wrap(~ dept) +
        theme(legend.position = 'bottom', 
              legend.title = element_blank()) +
        guides(fill = guide_legend(reverse = TRUE)) + 
        geom_vline(xintercept = 1.5, size = 0.5, linetype = 'dashed'))#, color = 'grey50')

# ggsave(filename = here('07_slides', '2021-05-14', 'images', 
#                        '5102_dept_bdo_level2.png'), 
#        plot = pl_5102_l2_bdo_stack, 
#        width = 10, 
#        height = 6, 
#        dpi = 125
#        )
```


```{r}
### Level 2 ethincity - by BDO & Class Groupings ----

##### Optional: enter dept name you'd like for the plot vs dept name listed 
##### in 5102 dataset
bdo_titles <- c("ARB" = 'Air Resources Board', 
                "OEHHA" = 'Environmental Health Hazard Assessment, Office of',
                "CalEPA (Agency)" = 'Environmental Protection Agency',
                "DPR" = 'Pesticide Regulation, Department of',
                "CalRecycle" = 'Resources Recycling and Recovery, Department of',
                "DTSC" = 'Toxic Substances Control, Department of',
                "SWRCB" = 'Water Resources Control Board')
bdo_titles_filenames <- tolower(c('ARB', 'OEHHA', 'CalEPA_Agency', 
                                  'DPR', 'CalRecycle', 'DTSC', 'SWRCB'))

# To get facet labels
pl_5102_groups_l2_bdo_all_facets <- map2_df(.x = names(bdo_titles), 
                                            .y = unname(bdo_titles), 
                                            .f = ~  df_5102_dept_1yr %>% 
                                                filter(dept == .y) %>% 
                                                fun_summary_5102_groups_l2() %>% 
                                                filter(type == 'Department Workforce') %>% 
                                                count(metrics_group, wt = ethnicity_total) %>% 
                                                mutate(facet_label = glue('{metrics_group} ({.x} n = {n})')) %>% 
                                                mutate(bdo = glue('{.x}'))
                                            )
```



```{r }
#### To make plots
pl_5102_groups_l2_bdo_all <- map2(.x = names(bdo_titles), 
                                  .y = unname(bdo_titles), 
                                  .f = ~ df_5102_dept_1yr %>% 
                                      filter(dept == .y) %>% 
                                      fun_summary_5102_groups_l2() %>%
                                      mutate(type = case_when(as.character(type) == 'Department Workforce' ~
                                                                  glue('{.x}'),
                                                              TRUE ~ as.character(type))) %>%
                                      mutate(type = factor(type, 
                                                           levels = c(.x, 'State Population'))) %>%
                                      mutate(bdo = glue('{.x}')) %>%
                                      left_join(pl_5102_groups_l2_bdo_all_facets,
                                                by = c('metrics_group', 'bdo')) %>%
                                      select(-metrics_group) %>%
                                      rename('metrics_group' = 'facet_label') %>%
                                      distinct() %>%
                                      fun_plot_5102_groups_l2_stack(plot_title = glue('{.x} Workforce by Occupation and Ethnicity Group vs. State Population (Year {report_year})')) %>%
                                      {.}
)

# ## View one plot
# pl_5102_groups_l2_bdo_all[[2]]

## print all plots
for (i in seq_along(pl_5102_groups_l2_bdo_all)) {
    print(pl_5102_groups_l2_bdo_all[[i]] + 
              theme(legend.position = 'bottom', 
                    legend.title = element_blank(),
                    plot.margin = margin(t = 30, r = 0, b = 30, l = 0, unit = 'pt') # adding a little bit of space between plots
                    )
          ) 
}

```

*To save these graphs as png image files, remove the comment symbols (#) from the* 
*code below.*

```{r}
# walk2(.x = pl_5102_groups_l2_bdo_all, 
#       # .y = tolower(str_replace_all(string = names(bdo_titles), 
#       #                              pattern = ' ', 
#       #                              replacement = '_') %>% 
#       #                  str_remove_all(pattern = '\\.|(|)')), 
#       .y = bdo_titles_filenames, # keep filenames consistent regardless of changes to plot labels/titles
#       .f = ~ ggsave(filename = here('07_slides', '2021-05-14', 'images', 
#                                     glue('5102_bdo_group_l2_{.y}.png')), 
#                     plot = .x,
#                     width = 10, 
#                     height = 6, 
#                     dpi = 125)
# )
```