---
title: "Data Prep"
author: "CalEPA Racial Equity Team"
date: '2022-03-18'
output: html_document
---

```{r setup, include=FALSE}
library(readxl)
library(readr)
library(dplyr)
library(janitor)
library(purrr)
library(glue)
library(lubridate)
library(stringr)
library(forcats)
library(scales)
library(tidycensus)
library(sf)

knitr::opts_chunk$set(echo = TRUE)
```


# Census Data (ACS 1 yr)
Define years
```{r}
years <- c(2011:2019, 2021)
```

API key
To download the census data, you need to first request a key here: <https://api.census.gov/data/key_signup.html/>
Then, enter census_api_key("*your_key*", install = TRUE) in your console.
```{r}
api_key <- "3e6e49b4c5dcaf87f3683870589946fa111e6852"
census_api_key(api_key, install=TRUE)
```


```{r}
# all years except 2020
acs_data_not2020 <- map_dfr(
    years,
    ~ get_acs(
        geography = 'state',
        variables = c(
            # total_state_pop = 'B02001_001',
            'Hispanic or Latino' = 'B03002_012',
            # Total Hispanic or Latino
            'White' = 'B03002_003',
            # White (Not Hispanic or Latino)
            'Black or African American' = 'B03002_004',
            # Black or African American (Not Hispanic or Latino)
            'Native American or Alaska Native' = 'B03002_005',
            # American Indian and Alaska Native (Not Hispanic or Latino)
            'Asian' = 'B03002_006',
            # Asian (Not Hispanic or Latino)
            'Pacific Islander' = 'B03002_007',
            # Native Hawaiian and Other Pacific Islander (Not Hispanic or Latino)
            'Other' = 'B03002_008',
            # Some other race (Not Hispanic or Latino)
            'Multiple' = 'B03002_009'# Two or more races (Not Hispanic or Latino)
        ),
        summary_var = c(total_state_pop = 'B02001_001'),
        survey = 'acs1',
        # use 'acs1' or 'acs5' to get 1 or 5 year acs
        state = 'CA',
        geometry = FALSE,
        # set to TRUE to get as geospatial data
        year = .x
    ),
    .id = "year"
)

# append 2020 data separately using 2019 ACS, consistent with R Markdown
acs_data_2020 <- get_acs(
        geography = 'state',
        variables = c(
            # total_state_pop = 'B02001_001',
            'Hispanic or Latino' = 'B03002_012',
            # Total Hispanic or Latino
            'White' = 'B03002_003',
            # White (Not Hispanic or Latino)
            'Black or African American' = 'B03002_004',
            # Black or African American (Not Hispanic or Latino)
            'Native American or Alaska Native' = 'B03002_005',
            # American Indian and Alaska Native (Not Hispanic or Latino)
            'Asian' = 'B03002_006',
            # Asian (Not Hispanic or Latino)
            'Pacific Islander' = 'B03002_007',
            # Native Hawaiian and Other Pacific Islander (Not Hispanic or Latino)
            'Other' = 'B03002_008',
            # Some other race (Not Hispanic or Latino)
            'Multiple' = 'B03002_009'# Two or more races (Not Hispanic or Latino)
        ),
        summary_var = c(total_state_pop = 'B02001_001'),
        survey = 'acs1',
        # use 'acs1' or 'acs5' to get 1 or 5 year acs
        state = 'CA',
        geometry = FALSE,
        # set to TRUE to get as geospatial data
        year = 2019
    )
```


Clean / reformat the ACS data (the data will be formatted to be consistent with the "level 2" ethnicity groupings in the workforce data that are created above)
```{r}
# First, modify years
acs_data_not2020_years <- acs_data_not2020 %>%
    mutate(year = as.numeric(year)) %>% 
    mutate(year = case_when(
        year < 10 ~ year + 2010,
        year >= 10 ~ year + 2011)) 

acs_data_2020_years <- acs_data_2020 %>% 
    mutate(year = 2020) 
    
# Bind 2020 with 2011-2019 and 2021 and modify column types & labels
acs_data_bind <- acs_data_not2020_years %>% 
    bind_rows(acs_data_2020_years) %>%
    clean_names() %>%
    rename(
        total_pop = summary_est,
        total_state_pop_moe = summary_moe,
        location_name = name,
        `Ethnicity` = variable
        )

# Group the 'Other' and 'Multiple' rows into one 'Other or Multiple Race' row
acs_data_level2 <- acs_data_bind %>%
    mutate(
        Ethnicity = case_when(
            Ethnicity == 'Other' | Ethnicity == 'Multiple' ~ 'Other or Multiple Race',
            TRUE ~ Ethnicity
        )
    ) %>%
    group_by(year, geoid, location_name, Ethnicity, total_pop) %>%
    summarize(ethnicity_total = sum(estimate)) %>%
    ungroup() %>%
    # Add a column with each ethnicity's % of total state population
    mutate(rate = ethnicity_total / total_pop) %>% 
    mutate(type = 'State Population',
           Level = "Level 2") %>%
    {.}

## create a dataset grouped at level 1 - all BIPOC together (resulting groups are BIPOC and white)
acs_data_level1 <- acs_data_bind %>%
    mutate(
        Ethnicity = case_when(
            Ethnicity == 'White' ~ 'White',
            TRUE ~ 'BIPOC')
    ) %>%
    group_by(year, geoid, location_name, Ethnicity, total_pop) %>%
    summarize(ethnicity_total = sum(estimate)) %>%
    ungroup() %>%
    mutate(rate = ethnicity_total / total_pop) %>% # update the rate to reflect level 1 numbers
    mutate(type = 'State Population',
           Level = "Level 1") %>% 
    {.}

acs_data_both_levels <- acs_data_level1 %>% 
    bind_rows(acs_data_level2)
```


## Export census data
```{r}
# setwd("05_shiny_app/data")
write.csv(acs_data_both_levels, file = "census_data.csv")
```



# 5102 Data
```{r}
url_data_all_yrs <- 'https://data.ca.gov/dataset/e620a64f-6b86-4ce0-ab4b-03d06674287b/resource/aba87ad9-f6b0-4a7e-a45e-d1452417eb7f/download/calhr_5102_statewide_2011-2021.csv'

df_5102_report <-
    read_csv(url_data_all_yrs,
             col_types = cols(.default = col_character()))

# mutate(record_count = as.integer(record_count)) %>%
# {
#     .
# }

# create a new column with just the year
df_5102_report_clean <- df_5102_report %>%
    type_convert() %>%
    clean_names() %>%
    mutate(report_year = year(as_of_date))


df_5012_level2 <- df_5102_report_clean %>%
    mutate(Ethnicity = case_when(
            str_detect(identity_variable, 'Pacific Islander') ~ 'Pacific Islander',
            # groups (all start with "Pacific Islander - "): Guamanian, Hawaiian, Other or Multiple, and Samoan
            str_detect(identity_variable, 'Asian') ~ 'Asian',
            # groups (all start with "Asian - "): Cambodian, Chinese, Filipino, Indian, Japanese, Korean, Laotian, Other or Multiple, Vietnamese
            TRUE ~ identity_variable
        )
    ) %>% 
    mutate(Level = "Level 2",
           type = 'Department Workforce')


df_5102_level1 <- df_5012_level2 %>% 
    mutate(Ethnicity = case_when(
            Ethnicity == 'White' ~ 'White',
            TRUE ~ 'BIPOC')) %>% 
    mutate(Level = "Level 1",
           type = 'Department Workforce')

workforce_data <- df_5102_level1 %>% 
    bind_rows(df_5012_level2)
```

## Export 5102 data
```{r}
# setwd("05_shiny_app/data")
write_csv(x = workforce_data, file = "workforce_data.csv.gz")
```


# Testing code for shiny app
TEST PAGE 3
```{r}
workforce_data_updated <- workforce_data %>%
    filter(dept == "Administrative Law, Office of",
           Level == "Level 1") %>%
    rename(year = report_year) %>%
    group_by(year) %>%
    add_count(Ethnicity,
              wt = record_count,
              name = 'ethnicity_total') %>%
    add_count(year,
              wt = record_count,
              name = 'total_pop') %>% 
    ungroup() %>% 
    mutate(rate = ethnicity_total / total_pop) %>%
    distinct(Ethnicity, year, .keep_all = TRUE) %>%
    arrange(Ethnicity) %>%
    select(year,
           Level,
           Ethnicity,
           total_pop,
           ethnicity_total,
           rate,
           type)
```


TEST PAGE 2
```{r}
 workforce_data_updated2 <- workforce_data %>%
    filter(dept == "Administrative Law, Office of",
           Level == "Level 1",
           report_year == 2017) %>%
    rename(year = report_year) %>%
    add_count(Ethnicity,
              wt = record_count,
              name = 'ethnicity_total') %>%
    distinct(Ethnicity, ethnicity_total, .keep_all = TRUE) %>%
    mutate(rate = ethnicity_total / sum(ethnicity_total)) %>%
    arrange(Ethnicity) %>%
    select(year,
           Level,
           Ethnicity,
           record_count,
           ethnicity_total,
           rate,
           type)
```


TEST PAGE 4
```{r}
categories <- workforce_data %>% 
    filter(Level == "Level 1",
           report_year == 2017) %>% 
    count(employee_category)

sub_categories <- workforce_data %>% 
    filter(Level == "Level 1",
           report_year == 2017) %>% 
    count(sub_category)

classes <- workforce_data %>% 
    filter(Level == "Level 1",
           report_year == 2017) %>% 
    count(class_title)



workforce_data_updated4 <- workforce_data %>%
    filter(#dept == "Administrative Law, Office of",
           Level == "Level 1",
           report_year == 2017) %>%
    rename(year = report_year) %>%
    add_count(Ethnicity,
              wt = record_count,
              name = 'ethnicity_total') %>%
    distinct(Ethnicity, ethnicity_total, .keep_all = TRUE) %>%
    mutate(rate = ethnicity_total / sum(ethnicity_total)) %>%
    arrange(Ethnicity) %>%
    select(year,
           Level,
           Ethnicity,
           record_count,
           ethnicity_total,
           rate,
           type)
```
